Notes: 

- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
- IRID039_S123 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings

to check SS27, SS01, IRID2,3,5-10, SS27

IRID025_S236 was eliminated


```{r}

pacman::p_load(googlesheets4, tidyverse, DT,lubridate, gtsummary, gt, rmarkdown)

# ----------------------------
# - Step 0: Adquiring metadata from google and do replacements in the strings
htlv_metadata = read_sheet("https://docs.google.com/spreadsheets/d/14eZCI8lRsfIebXu-tpAxbuPEOhE6moAnTDKKKUQyNOI/edit?gid=0#gid=0") |>
  mutate(ID=stringr::str_replace(ID, "^PE_", "")) |>
  mutate(ID=stringr::str_replace(ID, "_S.*", ""))
#----------------------------------------
# - Step 1: Manipulation 
htlv_mdata = htlv_metadata |>
  filter(Complete!="NO_HTLV") |>
  select(
    ID, 
    Status, 
    Birth_place_DNI, 
    Sex, 
    Age, 
    "PVL1 (copies/10000PMBCs)", 
    strongy_lifetime, 
    disease_2) |>
  dplyr::rename(
    PVL_home="PVL1 (copies/10000PMBCs)",
    birthplace=Birth_place_DNI
  ) |>
  mutate(
    Status = case_when(
      Status == "vivo" ~ "Alive",
      Status == "death" ~ "Death",
      .default=NULL
    ),
    id_birth = case_when(  
      toupper(birthplace) %in% c("AMAZONAS", "LORETO", "MADRE_DE_DIOS", "SAN_MARTIN", "UCAYALI") ~ "Amazonian region",
      toupper(birthplace)  %in% c("ANCASH", "AREQUIPA", "CALLAO", "ICA", "LA_LIBERTAD",   "LAMBAYEQUE", "LIMA", "MOQUEGUA", "PIURA", "TACNA", "TUMBES") ~ "Coastal region",
      toupper(birthplace) %in% c("PUNO", "HUANCAVELICA", "HUANUCO", "APURIMAC", "AYACUCHO", "CAJAMARCA", "JUNIN", "PASCO",  "CUSCO") ~ "Andean region"),
      Age=as.numeric(Age),
      Sex=recode(Sex, F = "Female", M = "Male"),
      disease_2=recode(disease_2, AC="Asymptomatic carriers", ATL="ATLL", HAM="HAM/TSP"),
      strongy_lifetime=recode(strongy_lifetime, YES="Strongyloides (+)",
      NO="Strongyloides (-)")

  )

# Preliminary review
datatable(htlv_mdata)
#---------------------------------------
# Table 1 - Final - gtsummary
table1 = htlv_mdata |>
  select(strongy_lifetime, -ID, -birthplace, -PVL_home, Sex, Age, id_birth, disease_2, Status, ) |>
  gtsummary::tbl_summary(
    by="strongy_lifetime",
    missing="no",
    type= list(
      Sex ~ "categorical",
      Age ~ "continuous",
      id_birth ~ "categorical",
      disease_2 ~ "categorical",
      Status ~ "categorical"),
    label=list(
      Sex ~ "Sex",
      Age ~ "Age(years)",
      id_birth ~ "Birthplace",
      disease_2 ~ "Disease",
      Status ~ "Mortality rate at 5 years"),
    ) |>
  gtsummary::add_overall(last=FALSE, col_label="**Total**") |>
  gtsummary::add_p() |>
  modify_header(label="**Variable**", all_stat_cols() ~ "**{level}**<br>N = {n} ({style_percent(p, digits=1)}%)") |>
  bold_labels() %>%
  as_gt()
table1
gtsave(table1, filename = "2503_Table1.docx", path="./output")
```

```{r}
pacman::p_load(tidyverse, ineq, tidyplots,DT)

#------------------
# Gini's index formula based on tags
gini_index <- function(counts) {
  n <- length(counts)
  mean_count <- mean(counts)
  sum_abs_diff <- sum(sapply(counts, function(x) sum(abs(x - counts))))
  gini <- sum_abs_diff / (2 * n^2 * mean_count)
  return(gini)
}

# Step 1 - Loading clonality total file
htlv_is = read_tsv("concatenated_clonality.txt", col_names = c("range", "chrom", "start", "end", "freq", "freq_p", "sample"))

# Option A for gini index
result <- htlv_is |>
  group_by(sample) |>
  summarise(gini = Gini(freq), .groups = 'drop') |>
  filter(gini0.4) # filtering those with high oligoclonality

# Option B - Official, using ineq and join
result <- htlv_is |>
group_by(sample) |>
group_split() |>
map_dfr(~tibble(
  sample = .x$sample[1], 
  gini = Gini(.x$freq),
  sum_freq = sum(.x$freq)
  )) |>
  mutate(
    sample=stringr::str_replace(sample, "^Peru_", "")) 

datatable(result)

# 1-----Plotting OCI vs Abundance (pearson)
result |>
    dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=gini, y=sum_freq, color=disease_2) |>
add_data_points(alpha=0.95) |>
adjust_x_axis(title = "Oligoclonality Index (OCI)", limits=c(0,1.1)) |>
adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance", limits=c(0,800)) |>
adjust_legend_title("Disease") |>
adjust_colors(new_colors = c("#3B78B0", "#E64B35B2","#7E6148B2")) |>
add_annotation_text(paste0("r = ", corr), x=0.25, y=700, fontsize = 7) |>
save_plot("output/oci.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 10,
  height = 7, 
  unit="cm")

corr=round(cor(result$gini, result$sum_freq, method = c("pearson")), 4)

datatable(result)
# 2-----Plotting OCI vs Strongy
result |>
 dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=strongy_lifetime, y=gini) |>
add_data_points_jitter(alpha=0.95) |>
add_violin(draw_quantiles = 0.5, alpha=0.2) |>
add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
adjust_y_axis(title = "Oligoclonality Index (OCI)", limits=c(-0.1,1.1)) |>
adjust_x_axis(title = NULL) |>
save_plot("output/S1_Strongi_vs_OCI.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 7,
  height = 8, 
  unit="cm") 

# 3-----Plotting OCI vs NGS reads
result |>
 dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=strongy_lifetime, y=sum_freq) |>
add_data_points_jitter(alpha=0.95) |>
add_violin(draw_quantiles = 0.5, alpha=0.2) |>
add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
adjust_x_axis(title = NULL) |>
save_plot("output/S1_Strongi_vs_NGSreads.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 7,
  height = 8, 
  unit="cm")
```


```{r}
"""
- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
IRID025_S236 was eliminated
to check SS27, SS01, IRID2,3,5-10

- IRID039_S123 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings
"""
pacman::p_load(tidyverse, midasHLA, styler, lintr, DT)
badquality = c("IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20")

# Step1: Filtering data, removing those without HTLV-1 clones or reads, and JP cases
htlv_la = read_tsv("concatenated_HLA.txt", col_names = c("none", "A_1", "A_2", "B_1", "B_2", "C_1", "C_2", "none_reads", "none_obj", "ID"))

htlv_hla = htlv_la|>
  select(-starts_with("none")) |>
  dplyr::filter(!grepl("ATLSkin|1339", ID)) |>
  dplyr::filter(!grepl(badquality, ID)) |>
  mutate(
    ID=stringr::str_replace(ID, "^Peru_", "")
    ) |>
    left_join(htlv_mdata, by=("ID")) |>
    left_join(result, by=c("ID"="sample"))

datatable(htlv_hla) #visualization
write_tsv(htlv_hla, "2503_filtered_HLA_rearranged.tsv")

# Step2: figures by 
A24A2 = htlv_hla |>
  mutate(
    A0201 = ifelse(grepl("A\\*02:01", A_1) | grepl("A\\*02:01", A_2), "A*02:01(+)", "A*02:01(-)"),
    A2402 = ifelse(grepl("A\\*24:02", A_1) | grepl("A\\*24:02", A_2),  "A*24:02(+)", "A*24:02(-)")) |>
    pivot_longer(cols=c(A0201, A2402), names_to="HLA", values_to="haplotype")|>
    mutate(HLA=recode(HLA, A0201 = "HLA-A*02:01", A2402="HLA-A*24:02"))

# comparison of IS by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_1_vs_NGSreads.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

    
# comparison of OCI by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_2_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", A_1) | grepl("B\\*35", A_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) #|>
    add_test_pvalue() |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) #|>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_2_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

show(fig_B35)

# Step2: read using midasHLA
pacman::p_load(midasHLA)

# Subsetting 
datatable(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)])

# Calling HLA and reducing it
hla_htlv = reduceHlaCalls(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)], resolution=4)

# Calculate HLA allele frequencies
getHlaFrequencies(
  hla_htlv,
  compare=TRUE,
  ref_pop=htlvpops,
  ref = allele_frequencies) |>
dplyr::filter(Freq > 0.05)
htlvpops[htlvpops %in% allele_frequencies$pop]
allele_frequencies$pop
htlvpops = c("Japan Okinawa Ryukyuan",
"Japan Central",
"USA NMDP Hispanic South or Central American",
"USA NMDP Japanese")
"Australia Yuendumu Aborigine",
"Papua New Guinea Wosera Abelam",
"Papua New Guinea Madang",
"Japan Okinawa Ryukyuan",
"Japan Central",
"China Zhejiang Han",
"Argentina Chiriguano",
"Peru Quechua",
"Peru Arequipa Mestizo",
"Gabon Haut-Ogooue Dienga",
"Brazil REDOME Bahia"
)
"Australia Yuendumu Aborigine" Indigenous Australian residents of central Australia 

"Papua New Guinea Wosera Abelam"
"Papua New Guinea Madang"
Japan pop 15
China Zhejiang Han
Argentina Chiriguano
Peru Quechua
Peru Arequipa Mestizo
Gabon Haut-Ogooue Dienga
Brazil REDOME Bahia

"Japan Okinawa Ryukyuan"
"Japan Central"
"USA NMDP Hispanic South or Central American"
"USA NMDP Japanese"
hla_htlv 
midas <- prepareMiDAS(
  hla_calls = hla_htlv,
  colData = htlv_clonality,
  experiment = "hla_alleles")

freq <- getFrequencies(
  object = midas,
  carrier_frequency = FALSE,
  experiment = "hla_alleles",
  compare = TRUE
)
freq |>
  filter(Counts>10)
midas_filtered_freq <-
  filterByFrequency(
    object = midas,
    experiment = "hla_alleles",
    lower_frequency_cutoff = 0.01
    )
midas_filtered_freq


object<-lm(gini~term, data=midas)
results <-
  runMiDAS(
    object = object,
    experiment = "hla_alleles",
    inheritance_model = "dominant",
    conditional = FALSE,
    omnibus = FALSE,
    lower_frequency_cutoff = 0.05
    )
kableResults(results)

htlv_clonality
results |>
  filter(p.value<0.05)


```

https://github.com/cancerit/dockstore-cgpmap/wiki/1.-Reference-files

ftp://ftp.sanger.ac.uk/pub/cancer/dockstore/human/GRCh38_hla_decoy_ebv/bwa_idx_GRCh38_hla_decoy_ebv_bwamem2.tar.gz
https://genentech.github.io/midasHLA/articles/MiDAS_vignette.html