Notes: 

- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
- IRID039_S123 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings

to check SS27, SS01, IRID2,3,5-10

IRID025_S236 was eliminated

```{r}

pacman::p_load(googlesheets4, tidyverse, DT,lubridate, gtsummary, gt, rmarkdown)

# ----------------------------
# - Step 0: Adquiring metadata from google and do replacements in the strings
htlv_metadata = read_sheet("https://docs.google.com/spreadsheets/d/14eZCI8lRsfIebXu-tpAxbuPEOhE6moAnTDKKKUQyNOI/edit?gid=0#gid=0") |>
  mutate(ID=stringr::str_replace(ID, "^PE_", "")) |>
  mutate(ID=stringr::str_replace(ID, "_S.*", ""))
#----------------------------------------
# - Step 1: Manipulation 
htlv_mdata = htlv_metadata |>
  filter(Complete!="NO_HTLV") |>
  select(
    ID, 
    Status, 
    Birth_place_DNI, 
    Sex, 
    Age, 
    "PVL1 (copies/10000PMBCs)", 
    strongy_lifetime, 
    disease_2) |>
  dplyr::rename(
    PVL_home="PVL1 (copies/10000PMBCs)",
    birthplace=Birth_place_DNI
  ) |>
  mutate(
    Status = case_when(
      Status == "vivo" ~ "Alive",
      Status == "death" ~ "Death",
      .default=NULL
    ),
    id_birth = case_when(  
      toupper(birthplace) %in% c("AMAZONAS", "LORETO", "MADRE_DE_DIOS", "SAN_MARTIN", "UCAYALI") ~ "Amazonian region",
      toupper(birthplace)  %in% c("ANCASH", "AREQUIPA", "CALLAO", "ICA", "LA_LIBERTAD",   "LAMBAYEQUE", "LIMA", "MOQUEGUA", "PIURA", "TACNA", "TUMBES") ~ "Coastal region",
      toupper(birthplace) %in% c("PUNO", "HUANCAVELICA", "HUANUCO", "APURIMAC", "AYACUCHO", "CAJAMARCA", "JUNIN", "PASCO",  "CUSCO") ~ "Andean region"),
      Age=as.numeric(Age),
      Sex=recode(Sex, F = "Female", M = "Male"),
      disease_2=recode(disease_2, AC="Asymptomatic carriers", ATL="ATLL", HAM="HAM/TSP"),
      strongy_lifetime=recode(strongy_lifetime, YES="Strongyloides (+)",
      NO="Strongyloides (-)")

  )

# Preliminary review
datatable(htlv_mdata)
#---------------------------------------
# Table 1 - Final - gtsummary
table1 = htlv_mdata |>
  select(strongy_lifetime, -ID, -birthplace, -PVL_home, Sex, Age, id_birth, disease_2, Status, ) |>
  gtsummary::tbl_summary(
    by="strongy_lifetime",
    missing="no",
    type= list(
      Sex ~ "categorical",
      Age ~ "continuous",
      id_birth ~ "categorical",
      disease_2 ~ "categorical",
      Status ~ "categorical"),
    label=list(
      Sex ~ "Sex",
      Age ~ "Age(years)",
      id_birth ~ "Birthplace",
      disease_2 ~ "Disease outcome",
      Status ~ "5-year mortality rate"),
    ) |>
  gtsummary::add_overall(last=FALSE, col_label="**Total**") |>
  gtsummary::add_p() |>
  modify_header(label="**Variable**", all_stat_cols() ~ "**{level}**<br>N = {n} ({style_percent(p, digits=1)}%)") |>
  bold_labels() %>%
  as_gt()
table1
gtsave(table1, filename = "2503_Table1.docx", path="./output")
```

```{r}
pacman::p_load(tidyverse, ineq, tidyplots,DT)

#------------------
# Gini's index formula based on tags
gini_index <- function(counts) {
  n <- length(counts)
  mean_count <- mean(counts)
  sum_abs_diff <- sum(sapply(counts, function(x) sum(abs(x - counts))))
  gini <- sum_abs_diff / (2 * n^2 * mean_count)
  return(gini)
}

# Step 1 - Loading clonality total file
htlv_is = read_tsv("concatenated_clonality.txt", col_names = c("range", "chrom", "start", "end", "freq", "freq_p", "sample"))

# Option A for gini index
result <- htlv_is |>
  group_by(sample) |>
  summarise(gini = Gini(freq), .groups = 'drop') |>
  filter(gini0.4) # filtering those with high oligoclonality

# Option B - Official, using ineq and join
result <- htlv_is |>
group_by(sample) |>
group_split() |>
map_dfr(~tibble(
  sample = .x$sample[1], 
  gini = Gini(.x$freq),
  sum_freq = sum(.x$freq)
  )) |>
  mutate(
    sample=stringr::str_replace(sample, "^Peru_", "")) 

datatable(result)

# 1-----Plotting OCI vs Abundance (pearson)
result |>
    dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=gini, y=sum_freq, color=disease_2) |>
add_data_points(alpha=0.95) |>
adjust_x_axis(title = "Oligoclonality Index (OCI)", limits=c(0,1.1)) |>
adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance", limits=c(0,800)) |>
adjust_legend_title("Disease") |>
adjust_colors(new_colors = c("#3B78B0", "#E64B35B2","#7E6148B2")) |>
add_annotation_text(paste0("r = ", corr), x=0.25, y=700, fontsize = 7) |>
save_plot("output/oci.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 10,
  height = 7, 
  unit="cm")

corr=round(cor(result$gini, result$sum_freq, method = c("pearson")), 4)

datatable(result)
# 2-----Plotting OCI vs Strongy
result |>
 dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=strongy_lifetime, y=gini) |>
add_data_points_jitter(alpha=0.95) |>
add_violin(draw_quantiles = 0.5, alpha=0.2) |>
add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
adjust_y_axis(title = "Oligoclonality Index (OCI)", limits=c(-0.1,1.1)) |>
adjust_x_axis(title = NULL) |>
save_plot("output/S1_Strongi_vs_OCI.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 7,
  height = 8, 
  unit="cm") 

# 3-----Plotting OCI vs NGS reads
result |>
 dplyr::right_join(htlv_mdata, by=c("sample"="ID")) |>
tidyplot(x=strongy_lifetime, y=sum_freq) |>
add_data_points_jitter(alpha=0.95) |>
add_violin(draw_quantiles = 0.5, alpha=0.2) |>
add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
adjust_x_axis(title = NULL) |>
save_plot("output/S1_Strongi_vs_NGSreads.png", 
  view_plot = TRUE,
  bg = "transparent",
  width = 7,
  height = 8, 
  unit="cm")
```


```{r}
"""
- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
IRID025_S236 was eliminated
to check SS27, SS01, IRID2,3,5-10

- IRID039_S123 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings
"""
pacman::p_load(tidyverse, midasHLA, styler, lintr, DT)
badquality = c("IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20")

# Step1: Filtering data, removing those without HTLV-1 clones or reads, and JP cases
htlv_la = read_tsv("concatenated_HLA.txt", col_names = c("none", "A_1", "A_2", "B_1", "B_2", "C_1", "C_2", "none_reads", "none_obj", "ID"))

htlv_hla = htlv_la|>
  select(-starts_with("none")) |>
  dplyr::filter(!grepl("ATLSkin|1339", ID)) |>
  dplyr::filter(!grepl(badquality, ID)) |>
  mutate(
    ID=stringr::str_replace(ID, "^Peru_", "")
    ) |>
    left_join(htlv_mdata, by=("ID")) |>
    left_join(result, by=c("ID"="sample"))

datatable(htlv_hla) #visualization
write_tsv(htlv_hla, "2503_filtered_HLA_rearranged.tsv")

# Step2: figures by 
A24A2 = htlv_hla |>
  mutate(
    A0201 = ifelse(grepl("A\\*02:01", A_1) | grepl("A\\*02:01", A_2), "A*02:01(+)", "A*02:01(-)"),
    A2402 = ifelse(grepl("A\\*24:02", A_1) | grepl("A\\*24:02", A_2),  "A*24:02(+)", "A*24:02(-)")) |>
    pivot_longer(cols=c(A0201, A2402), names_to="HLA", values_to="haplotype")|>
    mutate(HLA=recode(HLA, A0201 = "HLA-A*02:01", A2402="HLA-A*24:02"))

# comparison of IS by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_1_vs_NGSreads.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

    
# comparison of OCI by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_2_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35 and OCI
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35 part B
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_IS.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

show(fig_B35)

# Step2: read using midasHLA
pacman::p_load(midasHLA, tidyverse)


gra = hlaCallsGranthamDistance(
  hla_call,
  genes = "A",
  aa_selection = "binding_groove"
)
class(hla_call)
# Step3: Subsetting 
datatable(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)])

# Calling HLA and reducing it
hla_htlv = reduceHlaCalls(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)], resolution=4)

# Calculate HLA allele frequencies

comparison_hla = getHlaFrequencies(
  hla_htlv,
  compare=TRUE,
  ref = allele_frequencies) |>
  dplyr::filter(Freq > 0.05) 

write_tsv(comparison_hla, "output/2503_HLA_allelefrequencies.tsv")  

#Graphic >5% HLA frequencies
comparison_hla |>
dplyr::filter(Freq > 0.05) |>
  select(1,3,7,9,6,8,4) |>
  dplyr::rename(
    Latino="USA NMDP Hispanic South or Central American",
    Amerindian="USA NMDP North American Amerindian",
    Japanese="USA NMDP Japanese",
    European="USA NMDP European Caucasian",
    AfricanAmerican="USA NMDP African American pop 2"
    ) |>
    select(-Amerindian, -European)|>
    tidyr::pivot_longer(-allele,names_to="Population", values_to="haplotype_freq") |>
    mutate(Population=ifelse(Population=="Freq", "HTLV-1 Peru", Population)) |>
    mutate(Population=factor(Population, levels=c("HTLV-1 Peru", "Latino", "AfricanAmerican", "Japanese")))

population_colors=c("#C02D45", "#4FAE62", "#219ebc", "#ffb703")
 comparison_hla    |>
    tidyplot(x=haplotype_freq, y=allele, color=Population) |>
    add_mean_bar(width=.5) |>
    add_sum_dot() |>
    adjust_y_axis(title = "Alleles HLA-I > 5%") |>
    adjust_x_axis(title = "Haplotype Frequency", labels=scales::percent_format()) |>
    adjust_colors(new_colors = population_colors) |>
    save_plot("output/S3_HLAfreq.png", view_plot = TRUE,
    bg = "transparent",
    width = 10,
    height = 8, 
    unit="cm")
```


### Hardy-Weinberg-Equilibrium (HWE) to filter out alleles that strongly deviate from HWE expectations

The Hardy-Weinberg Equilibrium (HWE) is a fundamental principle in population genetics that describes the expected distribution of allele frequencies in a population under certain conditions (e.g., no mutation, migration, selection, or genetic drift). It states that allele and genotype frequencies in a population will remain constant from generation to generation in the absence of evolutionary influences.
```{r}
comparison_hla = getHlaFrequencies(
  hla_htlv,
  compare=TRUE,
  ref = allele_frequencies) |>
  dplyr::filter(Freq > 0.05) 
dim(comparison_hla)

custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome, conditional, omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
  HWE_cutoff = 0.05 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
  as.MiDAS = TRUE)

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    conditional=conditional,
    omnibus=omnibus,
    omnibus_groups_filter=additional
)
  return(HLA_results)
}
# metadata
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) #|>
  #dplyr::filter(!is.na(gini))
colhtlv$Status
# Adjust data if it includes 0s or 1s
#colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_df = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome)

# Define the models
models <- c("dominant", "recessive", "overdominant", "additive")
names(models) <- models  # Assign names to the models

# Initialize an empty list to store results
results_status <- list()
# Loop through the models
for (i in seq_along(models)) {
  # Apply custom_midas
  result <- custom_midas(
    hla_df = hla_df, 
    pheno_df = colhtlv, 
    exp = "hla_NK_ligands", 
    inheritance = models[i], 
    outcome = "Status",
    conditional=FALSE, #all results were not independent
    omnibus=FALSE,
    additional=NULL
  )
  
  # Add a column with the model name
  result$model <- names(models)[i]
  
  # Store the result in the list
  results_status[[i]] <- result
  combined_df <- do.call(rbind, results_status)
  return(combined_df)
}

datatable(combined_df)
write_tsv(combined_df, "output/2503_HLAanalysis_bydisease.tsv")  



````

```{r}
custom_midas_het <- function(hla_df, pheno_df, exp, outcome){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = c("hla_het", "hla_divergence"))

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    exponentiate = TRUE
)
  return(HLA_results)
}

result <- custom_midas_het(
    hla_df = hla_df, 
    pheno_df = colhtlv,
    outcome="disease_2",
    exp = "hla_divergence") 
result

```

```{r}
# FOR IS PVL 
custom_midas_pvl <- function(hla_df, pheno_df, exp, inheritance,outcome,omnibus){ #, conditional) #omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
    HWE_cutoff = 0.10 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
    as.MiDAS = TRUE)

  HLA_model <- glm.nb(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA)
    
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    #conditional=conditional,
    omnibus=omnibus,
    #omnibus_groups_filter=additional
)
  return(HLA_results)
}

##########
# data
# metadata
colhtlv2=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(sum_freq)) |>
  mutate(sum_freq=as.numeric(sum_freq))

hla_df2 = htlv_hla |>
  dplyr::filter(!is.na(sum_freq)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()
library(MASS)
models <- c("dominant", "recessive", "overdominant", "additive")
resultx = custom_midas_pvl(hla_df=hla_df2, pheno_df=colhtlv2, exp="hla_aa", inheritance="additive",outcome="sum_freq", omnibus=TRUE)
datatable(resultx)
```
```{r}
### ANALYSIS ASSOCIATION
#-- First, I am exploring the distribution of sum_freq (PVL) or gini index
#-----------------GINI INDEX (OCI) it is proportion (%)
htlv_hla <- htlv_hla[!is.na(htlv_hla$gini), ] #removing NAs

# Q-Q plot for normality (continuous data)
qqnorm(htlv_hla$gini)
qqline(htlv_hla$gini, col = "red") #it seems that it doesnt follow ND.

# Histogram of the response variable
hist(htlv_hla$gini, breaks = 30, main = "Distribution of Gini", xlab = "Gini")

"""
Beta regression model is designed for continuous data between 0 and 1. It uses a logit link function to model the mean of the Beta distribution
If the data is unimodal and symmetric, a Beta regression might be appropriate.
If the data is bimodal (peaks at 0 and 1), consider a zero-one-inflated Beta regression.
"""

# Simulate theoretical quantiles from the Beta distribution
fit_beta <- betareg(gini ~ 1, data = htlv_hla)
theoretical_quantiles <- qbeta(ppoints(length(htlv_hla$gini)), 
                               shape1 = fit_beta$coefficients$mean, 
                               shape2 = fit_beta$coefficients$precision)

# Create Q-Q plot
qqplot(theoretical_quantiles, htlv_hla$gini,
       main = "Q-Q Plot for Beta Distribution",
       xlab = "Theoretical Quantiles (Beta)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")
pacman::p_load(betareg)

# Corrected Q-Q plot for Poisson distribution
qqplot(qpois(ppoints(length(htlv_hla$gini)), lambda = mean(htlv_hla$gini)), 
       htlv_hla$gini,
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")



qqplot(qpois(ppoints(length(htlv_hla$sum_freq), lambda = mean(htlv_hla$sum_freq)), 
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")


 HLA = HWETest(
  object=midas,
  HWE_cutoff = 0.05 / 53,
  experiment = "hla_NK_ligands",
  as.MiDAS=TRUE)

# gini index as outcomd
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(gini))

# Adjust data if it includes 0s or 1s
colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_gini = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

# HLA divergence
midas <- prepareMiDAS(
  hla_calls = hla_gini,
  colData = colhtlv,
  experiment = "hla_divergence")

# Linear model
HLA_model <- lm(gini ~ term, data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence", 
  exponentiate = TRUE
)
# Beta reg model 
HLA_model <- gam(gini ~ term, family = betar(), data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence",
  inheritance_model=""
)

HLA_results

dispersion <- sum(residuals(HLA_model, type = "pearson")^2) / df.residual(HLA_model)
dispersion # much greater than 1, the data is overdispersed.



HLA
freq |>
  filter(Counts>10)
midas_filtered_freq <-
  filterByFrequency(
    object = midas,
    experiment = "hla_alleles",
    lower_frequency_cutoff = 0.01
    )
midas_filtered_freq


object<-lm(gini~term, data=midas)
results <-
  runMiDAS(
    object = object,
    experiment = "hla_alleles",
    inheritance_model = "dominant",
    conditional = FALSE,
    omnibus = FALSE,
    lower_frequency_cutoff = 0.05
    )
kableResults(results)

htlv_clonality
results |>
  filter(p.value<0.05)

htlv_hla
```

https://github.com/cancerit/dockstore-cgpmap/wiki/1.-Reference-files

ftp://ftp.sanger.ac.uk/pub/cancer/dockstore/human/GRCh38_hla_decoy_ebv/bwa_idx_GRCh38_hla_decoy_ebv_bwamem2.tar.gz
https://genentech.github.io/midasHLA/articles/MiDAS_vignette.html

https://hla.or.jp/med/frequency_search/en/allele/



## HLA-I or MHC-I
are divided into:
- Classical HLA-I: 
A, B, C. Immunosurveillance and cancer immunotherapy.
60-90% in cancer are affected. (immunescape)
a-heavy chain (HLA-A, HLA-B, HLA-C)
b-2-microglobulin.

After ubiquitinylation, proteins degraded by immunoproteasome.
Transportation by the transporter associated with antigen processing (TAP1/TAP2) into the ER.
Tapasin + chaperones (calreticulin, ERp57)
- Non-classical HLA-I: 

### 