Hypothesis: High sequence divergence between the alleles' peptide-binding domain strongly affects the combined peptide-binding properties.

Evolutionary divergence of the human leucocyte antigen class I (HLA-I) genes was evaluated using the Grantham distance at HLA-A, HLA-B, and HLA-C. Grantham distance takes into account composition, polarity, and volume.
HLA-B > HLA-A or HLA-C
Hierarchical clustering
```{r}
pacman::p_load(tidyverse, DT, tidyplots, httpgd, languageserver, midasHLA)
htlv_hla_la <- read_tsv("250304_HLAI/hla-la/hla_la_concatenated_output.txt") |>
    dplyr::filter(Q1 > .8)

# Ploting coverage
tidyplot(x = Locus, y = AverageCoverage, data = htlv_hla_la) |>
    adjust_x_axis(rotate_labels = TRUE) |>
    add_data_points_beeswarm(white_border = FALSE, size = 0.4, alpha = 0.4) |>
    adjust_y_axis_title("NGS Average Coverage") |>
    save_plot("output/S3_HLA_NGS_coverage.png",
        view_plot = TRUE,
        bg = "transparent",
        width = 8,
        height = 8,
        unit = "cm"
    )

datatable(htlv_hla_la)

# Wide format of HLA
htlv_hla <- htlv_hla_la |>
    select(1:3, 13) |>
    mutate(allele = paste0(Locus, "_", Chromosome)) |>
    dplyr::rename(ID = Sample) |>
    select(ID, allele, Allele) |>
    mutate(Allele = gsub("G$|N$", "", Allele)) |>
    pivot_wider(
        names_from = allele,
        id_cols = ID,
        values_from = Allele
    ) |>
    filter(!grepl("ATL|IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20", ID))

datatable(htlv_hla)

# HLA frequency calling

hla_htlv_call <- reduceHlaCalls(htlv_hla, resolution = 4)

getHlaFrequencies(
    hla_htlv_call,
    compare = TRUE,
    ref = allele_frequencies
) |>
    dplyr::filter(Freq > 0.01)

```


```{r}
library(tidyplots, DT, ggplot2)

plot_OR <- read_tsv("./output/2503_HLAanalysis_bydeath.tsv") |>
    dplyr::filter(p.value < 0.05) |>
    rename("CI_lower" = "conf.low", "CI_upper" = "conf.high") #|>

plot1 <- plot_OR |>
    tidyplot(x = estimate, y = allele) |>
    add_data_points() |>
    add(ggplot2::geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = .3, position = position_dodge(.9))) |>
    adjust_x_axis(
        title = "OR-adjusted to colinearity-LD",
        # limits=c(-5, 8)
    ) |>
    adjust_y_axis(title = "Alleles with p<0.05") |>
    add_annotation_line(x = 1, xend = 1.1, y = 0, yend = Inf, color = "grey") |>
    remove_legend() #|>

save_plot(plot1,
    filename = "./output/ORdeath.png",
    width = NA,
    height = NA,
    bg = "transparent"
)

datatable(plot_OR)
```


```{r}
## Evolutionary divergence for HLA 4 digits analysis
## -------------------------------------------------
pacman::p_load(tidyverse, tidyplots, DT, stringr, forcats)
### Loading data from optitype

htlv_hla <- read_tsv("2503_filtered_HLA_rearranged.tsv")

hla_call <- reduceHlaCalls(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)], resolution = 4)

# renaming for the HED in python
hla_call |>
    rename(
        "Sample" = "ID",
        "A1" = "A_1",
        "A2" = "A_2",
        "B1" = "B_1",
        "B2" = "B_2",
        "C1" = "C_1",
        "C2" = "C_2"
    ) |>
    write_tsv("2503_hla4dig.tsv")

# Loading HED distances
hed <- read_tsv("test.txt")
hed |>
    pivot_longer(-Sample, names_to = "HLA1", values_to = "HED") |>
    mutate(
        HLA1 = recode_factor(
            as.factor(HLA1),
            "HED_A" = "HLA-A",
            "HED_B" = "HLA-B",
            "HED_C" = "HLA-C",
            "Mean_HE" = "Mean HED",
            .ordered = TRUE
        )
    ) |>
    tidyplot(x = HLA1, y = HED, color = HLA1) |>
    add_violin() |>
    add_boxplot() |>
    adjust_x_axis_title("") |>
    remove_legend() |>
    add_test_asterisks()


# Heatmap of HLA-I
hed |>
    pivot_longer(-Sample, names_to = "HLA1", values_to = "HED") |>
    mutate(
        HLA1 = recode_factor(
            as.factor(HLA1),
            "HED_A" = "HLA-A",
            "HED_B" = "HLA-B",
            "HED_C" = "HLA-C",
            "Mean_HE" = "Mean HED",
            .ordered = TRUE
        )
    ) |>
    tidyplot(x = Sample, y = Sample, color = HED) |>
    add_heatmap()

datatable(hed)
hed
mean(hed$HED_C)
hed |>
    left_join(htlv_hla, by = c("Sample" = "ID")) |>
    tidyplots::tidyplot(y = Mean_HE, x = gini, color = disease_2) |>
    tidyplots::add_data_points()
fig
```

Using optitype, pairwise divergences of HLA-A, HLA-B, and HLA-C were similar


## Using HLA-LA files
```{r}
hlala_htlv <- read_tsv("250304_HLAI/hla-la/hla_la_concatenated_output.txt")

hlala_dataset <- hlala_htlv |>
    filter(Locus %in% c("A", "B", "C")) |>
    select(Locus, Chromosome, Allele, Sample) |>
    mutate(Allele = stringr::str_replace(Allele, "G", "")) |>
    mutate(Allele = str_trunc(Allele, width = 7, "right", ellipsis = "")) |>
    mutate(Alleles = paste0(Locus, Chromosome)) |>
    pivot_wider(values_from = "Allele", names_from = "Alleles", id_cols = "Sample") |>
    write_tsv("data/2503_hla_la_htlv.tsv")

datatable(hlala_dataset)
```

S20 seems to have a lower Quality for B

https://github.com/sunhuaibo/HLA-HED
```{python}
## hed to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -o data/2503_hla_la_hed.txt

## paired to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -p -o data/2503_hla_la_paired.txt
```


```{r}
read_tsv("data/2503_hla_la_hed.txt") |>
    mutate(country = ifelse(grepl("ATLSkin", Sample), "Japan", "Peru")) |>
    tidyplot(x = country, y = Mean_HE) |>
    add_boxplot() |>
    add_violin() |>
    add_test_pvalue()

wide_data <- read_tsv("data/2503_hla_la_paired.txt") |>
    pivot_wider(names_from = "Allele2", values_from = "HED")

wide_data
matrix_data_sorted <- wide_data |>
    arrange(Allele1) |>
    select(Allele1, sort(colnames(wide_data)[-1])) |>
    column_to_rownames("Allele1") |>
    as.matrix()

# Convert to a phylogenetic tree
library(ape)
tree <- njs(matrix_data_sorted) # Neighbor-joining tree
plot(tree)
matrix_data_sorted

# Heatmap
pairs1 <- read_tsv("data/2503_hla_la_paired.txt") |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    )
pairs1

# Create reverse pairs
reverse_pairs <- pairs1 |>
    mutate(Allele1A = Allele2, Allele2A = Allele1) |>
    filter(!paste(Allele1A, Allele2A) %in% paste(pairs1$Allele1, pairs1$Allele2)) |>
    mutate(Allele1 = Allele1A, Allele2 = Allele2A) |>
    select(-Allele1A, -Allele2A)

reverse_pairs

# Combine original data and reverse pairs
complete_data <- bind_rows(pairs1, reverse_pairs) |>
    arrange(desc(Allele1), Allele2) |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    )


complete_data |>
    tidyplot(x = Allele1, y = Allele2, color = HED) |>
    add_heatmap() |>
    adjust_colors(colors_continuous_plasma) |>
    add(ggplot2::facet_grid(rows = vars(A2), cols = vars(A1), scales = "free", space = "free", margins = FALSE, switch = "both")) |>
    remove_x_axis_labels() |>
    remove_y_axis_labels() |>
    remove_x_axis_title() |>
    remove_y_axis_title() |>
    remove_x_axis_ticks() |>
    remove_y_axis_ticks() |>
    save_plot(
        filename = "./output/2503_heatmap_hla.png",
        width = NA,
        height = NA,
        bg = "transparent"
    )


```


https://github.com/aiminy/IS-Seq-python3

https://github.com/nf-core/viralrecon

https://github.com/erickcastelli/HLA_genotyping

https://github.com/yshen25/HLA-Clus

https://www.youtube.com/watch?v=NNOEqMe4n74

https://grunwaldlab.github.io/Population_Genetics_in_R/

https://grunwaldlab.github.io/Population_Genetics_in_R/Data_Preparation.html

https://github.com/PapenfussLab/gridss

https://github.com/ICBI/viGEN

https://github.com/ClinicalGenomicsGBG/Viral_Integration_Pipeline

https://github.com/rosericazondekon/virusSeqPipeline

https://academic.oup.com/bioinformaticsadvances/article/4/1/vbae150/7815721

https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2020.587306/full

https://github.com/GIGA-AnimalGenomics-BLV/Public/tree/master/PIC

https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-6647-4

https://github.com/ding-lab/VirusIntegrate
https://github.com/SannaAb/Viral_Integration_Pipe

https://github.com/sara-javadzadeh/FastViFi

https://github.com/aehrc/isling

https://github.com/nf-core/viralintegration/blob/0.1.1/README.md

https://github.com/slowkow/hlabud

https://github.com/rbentham/HLAdivR/tree/48b2058f383ea75dad6e3c0f63bb1623ef5f0650

https://github.com/Genentech/midasHLA/blob/1a5bebfa67dd0f710a5ae5d4668fde9d3044e68a/NEWS.md?plain=1#L77

https://www.nature.com/articles/s41409-024-02341-z