
## Methods

Following reads extraction from the chromosome 6, coverage and quality score analyses were performed for all HLA regions. High-resolution HLA typing was carried out using OptiType (4-field resolution for HLA-I) and HLA-LA (6-field resolution for HLA-I/II). Allotypes failing quality thresholds (score <0.8 or average coverage <20) were discarded. Results from both bioinformatics tools were pairwise compared, and considering the average coverage for HLA-II, only HLA-I was considered to further analysis.

HLA class I allele frequencies were calculated after reduction to 4-field resolution using the *reduceHlaCalls* function from the R package midasHLA, frequencies for the Peruvian population were obtained using the function *getHlaFrequencies* and compared to the USA National Marrow Donor Program (NMDP) populations that were obtained from Allele Frequency Net Database (http://www.allelefrequencies.net/).



# Results

HLA-I allele frequencies 

The Peruvian HTLV-1 cohort displayed distinct HLA-I allele distribution patterns, with A02:01 being the most frequent allele at the HLA-A locus (44%), followed by A24:02 (17%) and A68:01 (7%). The HLA-B locus showed equal frequencies for B35:01 and B35:05 (12% each), with B48:01 appearing at 11%. At the HLA-C locus, C04:01 predominated (29%), followed by C01:02 (18%) and C*07:02 (13%). Comparative analysis with the NMDP database (Figure A) and an independent HTLV-1 cohort from Miyazaki, Japan (Figures B) revealed population-specific variations in HLA distribution. Additional distribution frequencies for HLA-I and HLA-II are presented in Supplementary table 1. 

Considering the highNext, we 

These differences may contribute to distinct patterns of HTLV-1 susceptibility or disease progression in the Peruvian population. Fewer dominant alleles may constrain immune responses to HTLV-1.

Suplementary Figure 1: NGS average coverage of HLA locus from the Peruvian HTLV-1 cohort and the Miyazaki-Japan HTLV-1 cohort.

Figure 1: Haplotype frequency for HLA-I alleles greater than 5% in the Peruvian HTLV-1 cohort compared to a) relevant populations from the NMDP database and b) the Miyazaki-Japan HTLV-1 cohort.

Pairwise divergence

We aimed to explore HLA evolutionary divergence between Japanese and Peruvian HTLV-1 cases, considering that alleles diversity peptide-binding strongly affects immune response.

was performed using OptiType and HLA-LA
Evolutionary divergence of the human leucocyte antigen class I (HLA-I) genes was evaluated using the Grantham distance at HLA-A, HLA-B, and HLA-C. Grantham distance takes into account composition, polarity, and volume.
HLA-B > HLA-A or HLA-C
Hierarchical clustering
```{r}
"""
- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
IRID025_S236 was eliminated
to check SS27, SS01, IRID2,3,5-10

- IRID039_S123 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings
"""
####################
# HLA_S1_Locus average coverage and quality
####################
#---Loading packages
pacman::p_load(tidyverse, DT, tidyplots, languageserver, midasHLA, stringr)
badquality = c("IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20")

#---Loading HLA coverage and data
htlv_hla_la <- read_tsv("250304_HLAI/hla-la/v2/hla_la_concatenated_output.txt") |>
    as.data.frame() |>
    dplyr::filter(Q1 > .8, AverageCoverage>20) |>
    mutate(country=ifelse(str_detect(Sample, "IRID|SS"), "Peru", "Japan")) |>
    dplyr::filter(!grepl(badquality, Sample))

#---HLA coverage - Ploting coverage
htlv_hla_la |>
tidyplot(x = Locus, y = AverageCoverage, color=country) |>
    adjust_x_axis(rotate_labels = TRUE) |>
    add_data_points_beeswarm(white_border = FALSE, size = 0.4, alpha = 0.4) |>
    adjust_y_axis(transform = "log10") |>
    adjust_y_axis_title("NGS Average Coverage", face="bold") |>
    add_median_bar(alpha=.3) |>
    adjust_x_axis_title("HLA Locus", face="bold") |>
    adjust_title("Average Coverage per HLA Locus", , face="bold") |>
    adjust_colors(
        new=c("#E69F00","#C02D45")) |>
    adjust_legend_title("Country", face="bold") |>
    adjust_legend_position("right") |>
    save_plot("output/S3_HLA_NGS_coverage.png",
        view_plot = TRUE,
        bg = "transparent"
    )
```

```{r}
####################
# HLA_Fig1 Allele frequencies
####################
pacman::p_load(flextable, officer)
#---Wide format of HLA
htlv_hla_la_wide <- htlv_hla_la |>
    select(1:3, 13, 14) |>
    mutate(allele = paste0(Locus, "_", Chromosome)) |>
    dplyr::rename(ID = Sample) |>
    select(ID, allele, Allele, country) |>
    mutate(Allele = gsub("G$|N$", "", Allele)) |>
    pivot_wider(
        names_from = allele,
        id_cols = c(ID, country),
        values_from = Allele)

#---Peru frequencies   
peru_hla = htlv_hla_la_wide |>
    filter(country=="Peru") |>
    select(-country)
#---Japan frequencies
japan_hla = htlv_hla_la_wide |>
    filter(country=="Japan") |>
    select(-country)
japan_hla
#---frequency calling
peru_hlacall <- reduceHlaCalls(peru_hla, resolution = 4)
japan_hlacall <- reduceHlaCalls(japan_hla, resolution = 4)
#---Japan calling
japan_hlafreq = getHlaFrequencies(
    japan_hlacall,
    compare = FALSE,
    ref = allele_frequencies
) |>
    dplyr::rename("Miyazaki - Japan Cohort"="Freq") |>
    select(-Counts) 

#---Peru calling
peru_hlafreq = getHlaFrequencies(
    peru_hlacall,
    compare = FALSE,
    ref = allele_frequencies
    ) |>
    dplyr::rename("Peru_Cohort"="Freq") |>
    select(-Counts) |>
    full_join(japan_hlafreq, by="allele") |>
    arrange(allele, Peru_Cohort)
peru_hlafreq
#---Table for supplementary files
ft <- flextable(peru_hlafreq) %>%
  theme_box() %>%                     
  set_caption(caption = "Allele Frequencies in HTLV-1 from Peru vs Japan") %>%
  autofit()                           
#---Save to Word
doc <- read_docx() %>%                
  body_add_flextable(ft) %>%          
  print(target = "output/STable_Allele_Frequencies.docx") 

#---Suplementary Figure 2 - Japan vs Peru allele frequencies more than 5%
peru_hlafreq |>
    dplyr::filter(Peru_Cohort> 0.05, grepl("^A|^B|^C", allele)) |>
    pivot_longer(-allele, names_to="Country", values_to="Frequency") |>
    mutate(Frequency=ifelse(is.na(Frequency), 0, Frequency),
        Country=dplyr::recode(Country, "Peru_Cohort"="Peruvian Cohort",
        "Miyazaki - Japan Cohort"="Miyazaki-Japanese Cohort"
        )) |>
    tidyplot(x=Frequency, y=allele, color=Country) |>
    add_mean_bar(width=.5) |>
    add_sum_dot() |>
    adjust_y_axis_title("Alleles HLA-I > 5%", face="bold") |>
    adjust_x_axis_title("Haplotype Frequency", face="bold") |>
    adjust_x_axis(labels=scales::percent_format()) |>
    adjust_title("HLA-I Allele Frequency by Country", , face="bold") |>
    adjust_colors(
        new=c("#E69F00","#C02D45")) |>
    adjust_legend_title("HTLV-1 Cohort", face="bold") |>
    adjust_legend_position("right") |>
    save_plot("output/S3_HLAfreq_peru_japan.png", view_plot = TRUE,
    bg = "transparent")

#---Suplementary Figure 3 - Japan vs Peru allele frequencies more than 5%
peru_hlafreq_world = getHlaFrequencies(
    peru_hlacall,
    compare = TRUE,
    ref = allele_frequencies
    )
#---HLA frequencies compared to other places
peru_hlafreq_world = peru_hlafreq_world |>
dplyr::filter(Freq > 0.05) |>
  select(1,3,7,9,6,8,4) |>
  dplyr::rename(
    "Latino NMDP"="USA NMDP Hispanic South or Central American",
    Amerindian="USA NMDP North American Amerindian",
    "Japanese NMDP"="USA NMDP Japanese",
    European="USA NMDP European Caucasian",
    "African-American NMDP"="USA NMDP African American pop 2"
    ) |>
    select(-Amerindian, -European)|>
    tidyr::pivot_longer(-allele,names_to="Population", values_to="haplotype_freq") |>
    mutate(Population=ifelse(Population=="Freq", "HTLV-1 Peru", Population)) |>
    mutate(Population=factor(Population, levels=c("HTLV-1 Peru", "Latino NMDP",  "Japanese NMDP", "African-American NMDP")))
population_colors=c("#C02D45", "#4FAE62",  "#ffb703", "#219ebc")
peru_hlafreq_world    |>
    dplyr::filter(grepl("^A|^B|^C", allele)) |>
    tidyplot(x=haplotype_freq, y=allele, color=Population) |>
    add_mean_bar(width=.5) |>
    add_sum_dot(size=.9) |>
    adjust_y_axis_title("Alleles HLA-I > 5%", face="bold") |>
    adjust_x_axis_title("Haplotype Frequency", face="bold") |>
    adjust_legend_title("Population", face="bold") |>
    adjust_x_axis(labels=scales::percent_format()) |>
    adjust_colors(new_colors = population_colors) |>
    save_plot("output/S3_HLAfreq_by_NMDP.png", view_plot = TRUE,
    bg = "transparent")
```

## Statistics

```{r}
```

```{r}
gra = hlaCallsGranthamDistance(
  hla_call,
  genes = "A",
  aa_selection = "binding_groove"
)
class(hla_call)
# Step3: Subsetting 
datatable(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)])

# Calling HLA and reducing it
hla_htlv = reduceHlaCalls(htlv_hla[, c(which(colnames(htlv_hla) == "ID"), 1:6)], resolution=4)




```


```{r}

pacman::p_load(tidyverse, midasHLA, styler, lintr, DT)
badquality = c("IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20")

# Step1: Filtering data, removing those without HTLV-1 clones or reads, and JP cases
htlv_la = read_tsv("concatenated_HLA.txt", col_names = c("none", "A_1", "A_2", "B_1", "B_2", "C_1", "C_2", "none_reads", "none_obj", "ID"))

htlv_hla = htlv_la|>
  select(-starts_with("none")) |>
  dplyr::filter(!grepl("ATLSkin|1339", ID)) |>
  dplyr::filter(!grepl(badquality, ID)) |>
  mutate(
    ID=stringr::str_replace(ID, "^Peru_", "")
    ) |>
    left_join(htlv_mdata, by=("ID")) |>
    left_join(result, by=c("ID"="sample"))

datatable(htlv_hla) #visualization
write_tsv(htlv_hla, "2503_filtered_HLA_rearranged.tsv")

# Step2: figures by 
A24A2 = htlv_hla |>
  mutate(
    A0201 = ifelse(grepl("A\\*02:01", A_1) | grepl("A\\*02:01", A_2), "A*02:01(+)", "A*02:01(-)"),
    A2402 = ifelse(grepl("A\\*24:02", A_1) | grepl("A\\*24:02", A_2),  "A*24:02(+)", "A*24:02(-)")) |>
    pivot_longer(cols=c(A0201, A2402), names_to="HLA", values_to="haplotype")|>
    mutate(HLA=recode(HLA, A0201 = "HLA-A*02:01", A2402="HLA-A*24:02"))

# comparison of IS by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_1_vs_NGSreads.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

    
# comparison of OCI by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_2_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35 and OCI
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35 part B
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_IS.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

show(fig_B35)

```


### Hardy-Weinberg-Equilibrium (HWE) to filter out alleles that strongly deviate from HWE expectations

The Hardy-Weinberg Equilibrium (HWE) is a fundamental principle in population genetics that describes the expected distribution of allele frequencies in a population under certain conditions (e.g., no mutation, migration, selection, or genetic drift). It states that allele and genotype frequencies in a population will remain constant from generation to generation in the absence of evolutionary influences.
```{r}
comparison_hla = getHlaFrequencies(
  hla_htlv,
  compare=TRUE,
  ref = allele_frequencies) |>
  dplyr::filter(Freq > 0.05) 
dim(comparison_hla)

custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome, conditional, omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
  HWE_cutoff = 0.05 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
  as.MiDAS = TRUE)

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    conditional=conditional,
    omnibus=omnibus,
    omnibus_groups_filter=additional
)
  return(HLA_results)
}
# metadata
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) #|>
  #dplyr::filter(!is.na(gini))
colhtlv$Status
# Adjust data if it includes 0s or 1s
#colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_df = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome)

# Define the models
models <- c("dominant", "recessive", "overdominant", "additive")
names(models) <- models  # Assign names to the models

# Initialize an empty list to store results
results_status <- list()
# Loop through the models
for (i in seq_along(models)) {
  # Apply custom_midas
  result <- custom_midas(
    hla_df = hla_df, 
    pheno_df = colhtlv, 
    exp = "hla_NK_ligands", 
    inheritance = models[i], 
    outcome = "Status",
    conditional=FALSE, #all results were not independent
    omnibus=FALSE,
    additional=NULL
  )
  
  # Add a column with the model name
  result$model <- names(models)[i]
  
  # Store the result in the list
  results_status[[i]] <- result
  combined_df <- do.call(rbind, results_status)
  return(combined_df)
}

datatable(combined_df)
write_tsv(combined_df, "output/2503_HLAanalysis_bydisease.tsv")  



````

```{r}
custom_midas_het <- function(hla_df, pheno_df, exp, outcome){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = c("hla_het", "hla_divergence"))

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    exponentiate = TRUE
)
  return(HLA_results)
}

result <- custom_midas_het(
    hla_df = hla_df, 
    pheno_df = colhtlv,
    outcome="disease_2",
    exp = "hla_divergence") 
result

```

```{r}
# FOR IS PVL 
custom_midas_pvl <- function(hla_df, pheno_df, exp, inheritance,outcome,omnibus){ #, conditional) #omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
    HWE_cutoff = 0.10 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
    as.MiDAS = TRUE)

  HLA_model <- glm.nb(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA)
    
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    #conditional=conditional,
    omnibus=omnibus,
    #omnibus_groups_filter=additional
)
  return(HLA_results)
}

##########
# data
# metadata
colhtlv2=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(sum_freq)) |>
  mutate(sum_freq=as.numeric(sum_freq))

hla_df2 = htlv_hla |>
  dplyr::filter(!is.na(sum_freq)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()
library(MASS)
models <- c("dominant", "recessive", "overdominant", "additive")
resultx = custom_midas_pvl(hla_df=hla_df2, pheno_df=colhtlv2, exp="hla_aa", inheritance="additive",outcome="sum_freq", omnibus=TRUE)
datatable(resultx)
```
```{r}
### ANALYSIS ASSOCIATION
#-- First, I am exploring the distribution of sum_freq (PVL) or gini index
#-----------------GINI INDEX (OCI) it is proportion (%)
htlv_hla <- htlv_hla[!is.na(htlv_hla$gini), ] #removing NAs

# Q-Q plot for normality (continuous data)
qqnorm(htlv_hla$gini)
qqline(htlv_hla$gini, col = "red") #it seems that it doesnt follow ND.

# Histogram of the response variable
hist(htlv_hla$gini, breaks = 30, main = "Distribution of Gini", xlab = "Gini")

"""
Beta regression model is designed for continuous data between 0 and 1. It uses a logit link function to model the mean of the Beta distribution
If the data is unimodal and symmetric, a Beta regression might be appropriate.
If the data is bimodal (peaks at 0 and 1), consider a zero-one-inflated Beta regression.
"""

# Simulate theoretical quantiles from the Beta distribution
fit_beta <- betareg(gini ~ 1, data = htlv_hla)
theoretical_quantiles <- qbeta(ppoints(length(htlv_hla$gini)), 
                               shape1 = fit_beta$coefficients$mean, 
                               shape2 = fit_beta$coefficients$precision)

# Create Q-Q plot
qqplot(theoretical_quantiles, htlv_hla$gini,
       main = "Q-Q Plot for Beta Distribution",
       xlab = "Theoretical Quantiles (Beta)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")
pacman::p_load(betareg)

# Corrected Q-Q plot for Poisson distribution
qqplot(qpois(ppoints(length(htlv_hla$gini)), lambda = mean(htlv_hla$gini)), 
       htlv_hla$gini,
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")



qqplot(qpois(ppoints(length(htlv_hla$sum_freq), lambda = mean(htlv_hla$sum_freq)), 
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")


 HLA = HWETest(
  object=midas,
  HWE_cutoff = 0.05 / 53,
  experiment = "hla_NK_ligands",
  as.MiDAS=TRUE)

# gini index as outcomd
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(gini))

# Adjust data if it includes 0s or 1s
colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_gini = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

# HLA divergence
midas <- prepareMiDAS(
  hla_calls = hla_gini,
  colData = colhtlv,
  experiment = "hla_divergence")

# Linear model
HLA_model <- lm(gini ~ term, data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence", 
  exponentiate = TRUE
)
# Beta reg model 
HLA_model <- gam(gini ~ term, family = betar(), data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence",
  inheritance_model=""
)

HLA_results

dispersion <- sum(residuals(HLA_model, type = "pearson")^2) / df.residual(HLA_model)
dispersion # much greater than 1, the data is overdispersed.



HLA
freq |>
  filter(Counts>10)
midas_filtered_freq <-
  filterByFrequency(
    object = midas,
    experiment = "hla_alleles",
    lower_frequency_cutoff = 0.01
    )
midas_filtered_freq


object<-lm(gini~term, data=midas)
results <-
  runMiDAS(
    object = object,
    experiment = "hla_alleles",
    inheritance_model = "dominant",
    conditional = FALSE,
    omnibus = FALSE,
    lower_frequency_cutoff = 0.05
    )
kableResults(results)

htlv_clonality
results |>
  filter(p.value<0.05)

htlv_hla
```

https://github.com/cancerit/dockstore-cgpmap/wiki/1.-Reference-files

ftp://ftp.sanger.ac.uk/pub/cancer/dockstore/human/GRCh38_hla_decoy_ebv/bwa_idx_GRCh38_hla_decoy_ebv_bwamem2.tar.gz
https://genentech.github.io/midasHLA/articles/MiDAS_vignette.html

https://hla.or.jp/med/frequency_search/en/allele/



## HLA-I or MHC-I
are divided into:
- Classical HLA-I: 
A, B, C. Immunosurveillance and cancer immunotherapy.
60-90% in cancer are affected. (immunescape)
a-heavy chain (HLA-A, HLA-B, HLA-C)
b-2-microglobulin.

After ubiquitinylation, proteins degraded by immunoproteasome.
Transportation by the transporter associated with antigen processing (TAP1/TAP2) into the ER.
Tapasin + chaperones (calreticulin, ERp57)
- Non-classical HLA-I: 

### 

### Subanalysis: including all HLA-LA from Japan

```{r}
#---packages
pacman::p_load(tidyverse, DT, tidyplots, httpgd, languageserver, midasHLA, stringr)

#---data loading
htlv_hla_la <- read_tsv("250304_HLAI/hla-la/v2/hla_la_concatenated_output.txt") |>
    dplyr::filter(Q1 > .8)
#---data wrangling
hlala_dataset <- htlv_hla_la |>
    filter(Locus %in% c("A", "B", "C")) |>
    select(Locus, Chromosome, Allele, Sample) |>
    mutate(Allele = stringr::str_replace(Allele, "G", "")) |>
    mutate(Allele = str_trunc(Allele, width = 7, "right", ellipsis = "")) |>
    mutate(Alleles = paste0(Locus, Chromosome)) |>
    pivot_wider(values_from = "Allele", names_from = "Alleles", id_cols = "Sample") |>
    write_tsv("data/2503_hla_la_htlv_peru_and_japan.tsv")

#===After reading the file with hla_hed.py
#---reading the output
hla_hed = read_tsv("data/2503_hla_total_hed.tsv")
#---data wrangling to make a new variable for countries
badquality = c("IRID011|IRID026|IRID038|IRID025|SS04|SS14|SS02|SS20")

hla_hed = hla_hed |>
    mutate(country=ifelse(str_detect(Sample, "IRID|SS"), "Peru", "Japan")) |>
    dplyr::filter(!grepl(badquality, Sample))

hla_hed |>
    pivot_longer(
        -c(country, Sample), 
        names_to="HLAHED", 
        values_to="HED"
        ) |>
    mutate(
        HLAHED=HLAHED |> fct_recode(
            "Mean HLA-I"="Mean_HE",
            "HLA-A"="HED_A",
            "HLA-B"="HED_B",
            "HLA-C"="HED_C"
            ) |>
            fct_relevel(
                "Mean HLA-I",
                "HLA-A",
                "HLA-B",
                "HLA-C"
            )
        ) |> 
    tidyplot(x=country, 
        y=HED, 
        color=country) |>
    add_data_points_jitter() |>
    add_test_pvalue(
        hide_info = TRUE,
        label.size = 6/ggplot2::.pt
        ) |>
    add_violin() |>
    remove_legend() |>
    adjust_x_axis(
        title="",
        labels = function(x) paste0(x, "\n(n = ", table(hla_hed$country)[x], ")")
        ) |>
    adjust_y_axis(limits = c(0,NA), padding = c(0.1, NA)) |>
    adjust_title(face="bold") |>
        adjust_colors(
        new=c("#E69F00","#D55E00")
    ) |>
    split_plot(
        by = HLAHED, 
        ncol = 4, 
        nrow = 1,
        guides="keep") |>
    save_plot(
        filename = "./output/2503_hla_la_japan_vs_peru.png",
        width = NA,
        height = NA,
        bg = "transparent"
    )


```

```{r}
library(tidyplots, DT, ggplot2)

plot_OR <- read_tsv("./output/2503_HLAanalysis_bydeath.tsv") |>
    dplyr::filter(p.value < 0.05) |>
    rename("CI_lower" = "conf.low", "CI_upper" = "conf.high") #|>

plot1 <- plot_OR |>
    tidyplot(x = estimate, y = allele) |>
    add_data_points() |>
    add(ggplot2::geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = .3, position = position_dodge(.9))) |>
    adjust_x_axis(
        title = "OR-adjusted to colinearity-LD",
        # limits=c(-5, 8)
    ) |>
    adjust_y_axis(title = "Alleles with p<0.05") |>
    add_annotation_line(x = 1, xend = 1.1, y = 0, yend = Inf, color = "grey") |>
    remove_legend() #|>

save_plot(plot1,
    filename = "./output/ORdeath.png",
    width = NA,
    height = NA,
    bg = "transparent"
)

datatable(plot_OR)
```


Using optitype, pairwise divergences of HLA-A, HLA-B, and HLA-C were similar

S20 seems to have a lower Quality for B

HED scores were calculated using a python script implemented in HLA-HED. [https://github.com/sunhuaibo/HLA-HED]

```{python}
## hed to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -o data/2503_hla_la_hed.txt

## paired to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -p -o data/2503_hla_la_paired.txt
```


Background 
The antigen presentation from HLA-I represent the most effective control system to mantain HTLV-1 infection  from HTLV-1 products
THe spectrum of antigens presented by HLA type 
## Heatmap of HLA-I evolutionary distances (HED) from Peruvian HTLV-1 cases

The heatmap shows z-score-normalized HED across HLA-I subtypes in HTLV-1 cases from Peru. The blue to red gradient indicates low to high HED between allele pairs, respectively.

```{r}
#---Supplementary heatmap distances
pairs1 <- read_tsv("data/2503_hla_la_paired.txt") |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    )
#---Create reverse pairs
reverse_pairs <- pairs1 |>
    mutate(Allele1A = Allele2, Allele2A = Allele1) |>
    filter(!paste(Allele1A, Allele2A) %in% paste(pairs1$Allele1, pairs1$Allele2)) |>
    mutate(Allele1 = Allele1A, Allele2 = Allele2A) |>
    select(-Allele1A, -Allele2A)
#---Combine original data and reverse pairs
complete_data <- bind_rows(pairs1, reverse_pairs) |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    ) |>
    arrange(Allele1, Allele2)
#---Heatmap for supplementary data
complete_data |>
    tidyplot(x = Allele1, y = Allele2, color = HED) |>
    add_heatmap() |>
    adjust_colors(colors_diverging_spectral) |>
    add(ggplot2::facet_grid(rows = vars(A2), cols = vars(A1), scales = "free", space = "free", margins = FALSE, switch = "both")) |>
    reverse_y_axis_labels() |>
    remove_x_axis_labels() |>
    remove_y_axis_labels() |>
    remove_x_axis_title() |>
    remove_y_axis_title() |>
    remove_x_axis_ticks() |>
    remove_y_axis_ticks() |>
    add(theme(panel.spacing = unit(0.05, "lines"))) |>
    save_plot(
        filename = "./output/2503_heatmap_hla.png",
        width = NA,
        height = NA,
        bg = "transparent"
    )
```


https://github.com/aiminy/IS-Seq-python3

https://github.com/nf-core/viralrecon

https://github.com/erickcastelli/HLA_genotyping

https://github.com/yshen25/HLA-Clus

https://www.youtube.com/watch?v=NNOEqMe4n74

https://grunwaldlab.github.io/Population_Genetics_in_R/

https://grunwaldlab.github.io/Population_Genetics_in_R/Data_Preparation.html

https://github.com/PapenfussLab/gridss

https://github.com/ICBI/viGEN

https://github.com/ClinicalGenomicsGBG/Viral_Integration_Pipeline

https://github.com/rosericazondekon/virusSeqPipeline

https://academic.oup.com/bioinformaticsadvances/article/4/1/vbae150/7815721

https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2020.587306/full

https://github.com/GIGA-AnimalGenomics-BLV/Public/tree/master/PIC

https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-6647-4

https://github.com/ding-lab/VirusIntegrate
https://github.com/SannaAb/Viral_Integration_Pipe

https://github.com/sara-javadzadeh/FastViFi

https://github.com/aehrc/isling

https://github.com/nf-core/viralintegration/blob/0.1.1/README.md

https://github.com/slowkow/hlabud

https://github.com/rbentham/HLAdivR/tree/48b2058f383ea75dad6e3c0f63bb1623ef5f0650

https://github.com/Genentech/midasHLA/blob/1a5bebfa67dd0f710a5ae5d4668fde9d3044e68a/NEWS.md?plain=1#L77

https://www.nature.com/articles/s41409-024-02341-z

https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html