---
title: "Flowchart Tutorial"
author: "Daniel Enriquez"
date: "2023-11-22"
output: html_document
---
https://www.r-bloggers.com/2022/06/creating-flowcharts-with-ggplot2/
https://mikeyharper.uk/flowcharts-in-r-using-diagrammer/
https://epirhandbook.com/en/diagrams-and-charts.html

## Flowchart by Gmisc and grid
References:
https://cran.r-project.org/web/packages/Gmisc/vignettes/Grid-based_flowcharts.html


### * Requirements

```{r setup, include=FALSE}
library(pacman)
p_load(Gmisc,glue,htmlTable,grid,magrittr)
knitr::opts_chunk$set(echo = FALSE)
```

### * Basic components

 - Boxes: `boxGrob` and `boxPropGrob` functions.
 - Arrows: `connectGrob` function.
 - Spread position: `spreadHorizontal` and `spreadVertical` functions.
 - Alignment of boxes before or after spreading: `alignHorizontal` and `alignVertical`
 
#### * A simple box

```{r box}
grid.newpage()
box<-"All patients were
included for this study
-A new line
-Another line"
#It respects our breaklines or use \n
boxGrob(box)
```

#### * A box with style

```{r style}
grid.newpage()
boxGrob("Hola\nyo hablo\nnihongo\nok", #it has to include \n to break a line
        width = .8, height = .8,
        x = 0, y = 0,
        bjust = c("left", "bottom"),
        txt_gp = gpar(col = "darkblue", cex = 2),
        box_gp = gpar(fill = "transparent", col = "darkblue"))
```

#### * A box with proportions

```{r proportions}
grid.newpage()
boxPropGrob("This is how i am writing this papper", "Uncompleted", "Completed", prop=0.3)
```

#### * A box with coordinates

```{r coordinates}
grid.newpage()
smpl_bx <- boxGrob(
  label = "A simple box",
  x = .5,
  y = .9,
  just = "center")

prop_bx <- boxPropGrob(
  label = "A split box",
  label_left = "Left side",
  label_right = "Right side",
  x = .5,
  y = .3,
  prop = .3,
  just = "center")

plot(smpl_bx)
plot(prop_bx)

smpl_bx_coords <- coords(smpl_bx)
grid.circle(y = smpl_bx_coords$y,
            x = smpl_bx_coords$x,
            r = unit(2, "mm"),
            gp = gpar(fill = "#FFFFFF99", col = "black"))
grid.circle(y = smpl_bx_coords$bottom,
            x = smpl_bx_coords$right,
            r = unit(1, "mm"),
            gp = gpar(fill = "red"))
grid.circle(y = smpl_bx_coords$top,
            x = smpl_bx_coords$right,
            r = unit(1, "mm"),
            gp = gpar(fill = "purple"))
grid.circle(y = smpl_bx_coords$bottom,
            x = smpl_bx_coords$left,
            r = unit(1, "mm"),
            gp = gpar(fill = "blue"))
grid.circle(y = smpl_bx_coords$top,
            x = smpl_bx_coords$left,
            r = unit(1, "mm"),
            gp = gpar(fill = "orange"))

prop_bx_coords <- coords(prop_bx)
grid.circle(y = prop_bx_coords$y,
            x = prop_bx_coords$x,
            r = unit(2, "mm"),
            gp = gpar(fill = "#FFFFFF99", col = "black"))
grid.circle(y = prop_bx_coords$bottom,
            x = prop_bx_coords$right_x,
            r = unit(1, "mm"),
            gp = gpar(fill = "red"))
grid.circle(y = prop_bx_coords$top,
            x = prop_bx_coords$right_x,
            r = unit(1, "mm"),
            gp = gpar(fill = "purple"))
grid.circle(y = prop_bx_coords$bottom,
            x = prop_bx_coords$left_x,
            r = unit(1, "mm"),
            gp = gpar(fill = "blue"))
grid.circle(y = prop_bx_coords$top,
            x = prop_bx_coords$left_x,
            r = unit(1, "mm"),
            gp = gpar(fill = "orange"))

grid.circle(y = prop_bx_coords$bottom,
            x = prop_bx_coords$right,
            r = unit(2, "mm"),
            gp = gpar(fill = "red"))
grid.circle(y = prop_bx_coords$top,
            x = prop_bx_coords$right,
            r = unit(2, "mm"),
            gp = gpar(fill = "purple"))
grid.circle(y = prop_bx_coords$bottom,
            x = prop_bx_coords$left,
            r = unit(2, "mm"),
            gp = gpar(fill = "blue"))
grid.circle(y = prop_bx_coords$top,
            x = prop_bx_coords$left,
            r = unit(2, "mm"),
            gp = gpar(fill = "orange"))

```

#### * Connecting the boxes

```{r connectors}
grid.newpage()
# Initiate the boxes that we want to connect
side <- boxPropGrob("Side", "Left", "Right",
                    prop = .3,
                    x = 0, y = .9,
                    bjust = c(0,1))
start <- boxGrob("Top",
                 x = .6, y = coords(side)$y,
                 box_gp = gpar(fill = "yellow"))
bottom <- boxGrob("Bottom",
                  x = .6, y = 0,
                  bjust = "bottom")


sub_side_left <- boxGrob("Left",
                         x = coords(side)$left_x,
                         y = 0,
                         bjust = "bottom")
sub_side_right <- boxGrob("Right",
                          x = coords(side)$right_x,
                          y = 0,
                          bjust = "bottom")

odd <- boxGrob("Odd\nbox",
               x = coords(side)$right,
               y = .5)

odd2 <- boxGrob("Also odd",
               x = coords(odd)$right +
                 distance(bottom, odd, type = "h", half = TRUE) -
                 unit(2, "mm"),
               y = 0,
               bjust = c(1,0))

exclude <- boxGrob("Exclude:\n - Too sick\n - Prev. surgery",
                   x = 1,
                   y = coords(bottom)$top +
                     distance(start, bottom, type = "v", half = TRUE),
                   just = "left", bjust = "right")

# Connect the boxes and print/plot them
connectGrob(start, bottom, "vertical")
connectGrob(start, side, "horizontal")
connectGrob(bottom, odd, "Z", "l")
connectGrob(odd, odd2, "N", "l")
connectGrob(side, sub_side_left, "v", "l")
connectGrob(side, sub_side_right, "v", "r")
connectGrob(start, exclude, "-",
            lty_gp = gpar(lwd = 2, col = "darkred", fill = "darkred"))

# Print the grobs
start
bottom
side
exclude
sub_side_left
sub_side_right
odd
odd2

```

### * Basic diagrams for paper
https://bookdown.org/yihui/rmarkdown-cookbook/diagrams.html
```{r setup, include=FALSE}
#library(pacman)
#p_load(DiagrammeR)
pacman::p_load(DiagrammeR, networkD3, tidyverse)
#import data with rio package
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```


Basic diagram
```{r}
DiagrammeR::grViz("digraph {
  graph [layout = dot, rankdir = TB]
  
  node [shape = rectangle]        
  rec1 [label = 'Step 1. Wake up']
  rec2 [label = 'Step 2. Write code']
  rec3 [label =  'Step 3. ???']
  rec4 [label = 'Step 4. PROFIT']
  
  # edge definitions with the node IDs
  rec1 -> rec3 
  rec2 -> rec4
  }",
  height = 500)
```

Using Gmisc

```{r}
pacman::p_load(Gmisc, glue, htmlTable, grid, magrittr,extrafont)
```


example
```{r}
grid.newpage()

boxh1 <- boxGrob("A cool\nreference\nbox",
               x = .5, y = .8,
               box_gp = gpar(fill = "#ADB5C7"))
another_box <- boxGrob("A horizontal box", x = .1, y =  .5)
yet_another_box <- boxGrob("Another horizontal box", x = .8, y = .3)

alignedBoxes <- alignHorizontal(box,
                                another_box,
                                yet_another_box,
                                .position = "center")

box
for (b in alignedBoxes) {
  print(b)}
```

```{r}
library(grid)
grid.newpage()

box1 <- boxGrob("B1", x = .2, y = .8)
box2 <- boxGrob("B2", x = .2, y = .8)
box3 <- boxGrob("B3", x = .2, y = .8)
box4 <- boxGrob("B4", x = .2, y = .8)
box5 <- boxGrob("B5", x = .2, y = .8)

spread_boxes <- spreadVertical(box1,
                               box2,
                               box3,
                               box4,
                               box5, 
                               .type = "between")
for (b in spread_boxes) {
  print(b)
}

```


```{R}

```


```