```{r}
"""
- IRID011, IRID026 and IRID038 were excluded (no HTLV-1 clones detected)
IRID025_S236 was eliminated
to check SS27, SS01, IRID2,3,5-10

- IRID039 and SS04 (are the same)  confirmed
- IRID015 and SS14 (are the same) confirmed
- IRID062 and SS02 (are the same) confirmed
- IRID089 and SS20 (are the same) confirmed
- IRID015 and 032 are siblings
"""
####################
# HLA_S1_Locus average coverage and quality
####################

#---Loading packages
pacman::p_load(tidyverse, DT, tidyplots, languageserver, midasHLA, stringr)
pacman::p_load(flextable, officer)
serialsamples = c("IRID039|IRID015|IRID062|IRID089")

#---Loading HLA coverage and data
htlv_hla_la <- read_tsv("250304_HLAI/hla-la/v2/hla_la_concatenated_output.txt") |>
    as.data.frame() |>
    dplyr::filter(Q1 > .8, AverageCoverage>20) |>
    mutate(country=ifelse(str_detect(Sample, "IRID|SS"), "Peru", "Japan")) |>
    dplyr::filter(!grepl(serialsamples, Sample))

htlv_hla_la = htlv_hla_la|> filter(country=="Peru")
htlv_hla_la
DT::datatable(htlv_hla_la)
#---HLA coverage - Ploting coverage
htlv_hla_la |>
tidyplot(x = Locus, y = AverageCoverage) |>
    adjust_x_axis(rotate_labels = TRUE) |>
    add_data_points_beeswarm(white_border = FALSE, size = 0.4, alpha = 0.4) |>
    adjust_y_axis(transform = "log10") |>
    adjust_y_axis_title("NGS Average Coverage", face="bold") |>
    add_median_bar(alpha=.3) |>
    adjust_x_axis_title("HLA Locus", face="bold") |>
    adjust_title("Average Coverage per HLA Locus", , face="bold") |>
    adjust_colors(
        new=c("#E69F00")) |>
    save_plot("output/2510_S3_HLA_NGS_coverage.png",
        view_plot = TRUE,
        bg = "transparent"
    )
```

DT::datatable(htlv_hla_la)
#---HLA coverage - Ploting coverage
htlv_hla_la |>
tidyplot(x = Locus, y = AverageCoverage, color=country) |>
    adjust_x_axis(rotate_labels = TRUE) |>
    add_data_points_beeswarm(white_border = FALSE, size = 0.4, alpha = 0.4) |>
    adjust_y_axis(transform = "log10") |>
    adjust_y_axis_title("NGS Average Coverage", face="bold") |>
    add_median_bar(alpha=.3) |>
    adjust_x_axis_title("HLA Locus", face="bold") |>
    adjust_title("Average Coverage per HLA Locus", , face="bold") |>
    adjust_colors(
        new=c("#E69F00","#C02D45")) |>
    adjust_legend_title("Country", face="bold") |>
    adjust_legend_position("right") |>
    save_plot("output/S3_HLA_NGS_coverage.png",
        view_plot = TRUE,
        bg = "transparent"
    )
```

```{r}
####################
# HLA_Fig1 Allele frequencies
####################

#---Wide format of HLA
htlv_hla_la_wide <- htlv_hla_la |>
    dplyr::select(1:3, 13, 14) |>
    mutate(allele = paste0(Locus, "_", Chromosome)) |>
    dplyr::rename(ID = Sample) |>
    dplyr::select(ID, allele, Allele, country) |>
    mutate(Allele = gsub("G$|N$", "", Allele)) |>
    pivot_wider(
        names_from = allele,
        id_cols = c(ID, country),
        values_from = Allele)
DT::datatable(htlv_hla_la_wide)
#---Peru frequencies   
peru_hla = htlv_hla_la_wide |>
    filter(country=="Peru") |>
    dplyr::select(-country)
DT::datatable(peru_hla)
#---Japan frequencies
japan_hla = htlv_hla_la_wide |>
    filter(country=="Japan") |>
    dplyr::select(-country)
japan_hla
#---frequency calling
peru_hla2 = peru_hla |>
  dplyr::select(1:13)
peru_hlacall <- reduceHlaCalls(peru_hla2, resolution = 4)



japan_hlacall <- reduceHlaCalls(japan_hla, resolution = 4)
#---Japan calling
japan_hlafreq = getHlaFrequencies(
    japan_hlacall,
    compare = FALSE,
    ref = allele_frequencies
) |>
    dplyr::rename("Miyazaki - Japan Cohort"="Freq") |>
    dplyr::select(-Counts) 
peru_hlacall

#---Peru calling
peru_hlafreq = getHlaFrequencies(
    peru_hlacall,
    compare = FALSE,
    ref = allele_frequencies
    ) |>
    dplyr::rename("Peru_Cohort"="Freq") |>
    dplyr::select(-Counts) # |>
    full_join(japan_hlafreq, by="allele") |>
    arrange(allele, Peru_Cohort)

peru_hlafreq
#---Table for supplementary files
ft <- flextable(peru_hlafreq) %>%
  theme_box() %>%                     
  set_caption(caption = "Allele Frequencies in HTLV-1 from Peru vs Japan") %>%
  autofit()                           
#---Save to Word
doc <- read_docx() %>%                
  body_add_flextable(ft) %>%          
  print(target = "output/STable_Allele_Frequencies.docx") 

#---Suplementary Figure 2 - Japan vs Peru allele frequencies more than 5%
peru_hlafreq |>
    dplyr::filter(Peru_Cohort> 0.05, grepl("^A|^B|^C", allele)) |>
    pivot_longer(-allele, names_to="Country", values_to="Frequency") |>
    mutate(Frequency=ifelse(is.na(Frequency), 0, Frequency),
        Country=dplyr::recode(Country, "Peru_Cohort"="Peruvian Cohort",
        "Miyazaki - Japan Cohort"="Miyazaki-Japanese Cohort"
        )) |>
    tidyplot(x=Frequency, y=allele, color=Country) |>
    add_mean_bar(width=.5) |>
    add_sum_dot() |>
    adjust_y_axis_title("Alleles HLA-I > 5%", face="bold") |>
    adjust_x_axis_title("Haplotype Frequency", face="bold") |>
    adjust_x_axis(labels=scales::percent_format()) |>
    adjust_title("HLA-I Allele Frequency by Country", , face="bold") |>
    adjust_colors(
        new=c("#E69F00","#C02D45")) |>
    adjust_legend_title("HTLV-1 Cohort", face="bold") |>
    adjust_legend_position("right") |>
    save_plot("output/S3_HLAfreq_peru_japan.png", view_plot = TRUE,
    bg = "transparent")

#---Suplementary Figure 3 - Japan vs Peru allele frequencies more than 5%
peru_hlafreq_world = getHlaFrequencies(
    peru_hlacall,
    compare = TRUE,
    ref = allele_frequencies
    )
#---HLA frequencies compared to other places
peru_hlafreq_world = peru_hlafreq_world |>
dplyr::filter(Freq > 0.05) |>
  dplyr::select(1,3,7,9,6,8,4) |>
  dplyr::rename(
    "Latino NMDP"="USA NMDP Hispanic South or Central American",
    Amerindian="USA NMDP North American Amerindian",
    "Japanese NMDP"="USA NMDP Japanese",
    European="USA NMDP European Caucasian",
    "African-American NMDP"="USA NMDP African American pop 2"
    ) |>
    dplyr::select(-Amerindian, -European)|>
    tidyr::pivot_longer(-allele,names_to="Population", values_to="haplotype_freq") |>
    mutate(Population=ifelse(Population=="Freq", "HTLV-1 Peru", Population)) |>
    mutate(Population=factor(Population, levels=c("HTLV-1 Peru", "Latino NMDP",  "Japanese NMDP", "African-American NMDP")))
population_colors=c("#C02D45", "#4FAE62",  "#ffb703", "#219ebc")
datax =peru_hlafreq_world    |>
    dplyr::filter(grepl("^A|^B|^C", allele)) 
plotx = tidyplot(x=haplotype_freq, y=allele, color=Population, data=datax) |>
    add_mean_bar(width=.5) |>
    add_sum_dot(size=.9) |>
    adjust_y_axis_title("Alleles HLA-I > 5%", face="bold") |>
    adjust_x_axis_title("Haplotype Frequency", face="bold") |>
    adjust_legend_title("Population", face="bold") |>
    adjust_x_axis(labels=scales::percent_format()) |>
    adjust_colors(new_colors = population_colors) #|>
    save_plot("output/2510_HLAfreq_by_NMDP.tiff", view_plot = TRUE,
    bg = "transparent", )


ggsave("2510_hla_v2.tiff", plot = plotx,  path=here("output/"), scale=1, dpi=600, limitsize=FALSE)
```

## Statistics

```{r}
# Packages
pacman::p_load(adegenet, pegas, hierfstat)

# Reducing manually to 2-fields
htlv_df = htlv_hla_la_wide |>
  dplyr::select(1:8) |>
  mutate(across(
    .cols = 3:8, 
    .fns = ~ str_trunc(., width = 7, side = "right", ellipsis = "")
  ),
  HLA_A = paste(A_1, A_2, sep = "/"),
  HLA_B = paste(B_1, B_2, sep = "/"),
  HLA_C = paste(C_1, C_2, sep = "/"),
  ) |>
  dplyr::select(ID, country, HLA_A, HLA_B, HLA_C ) 


genind_obj <- df2genind(
  X = htlv_df[, c("HLA_A", "HLA_B", "HLA_C")],  
  sep = "/",                          
  ind.names = htlv_df$ID,        
  pop = htlv_df$country,            
  ploidy = 2                          
)
# Hardy-Weinberg Equilibrium (HWE)
hwe_test <- hw.test(genind_obj, B = 1000)
print(hwe_test)

# Weir & Cockerham’s FST
hierfstat_data <- genind2hierfstat(genind_obj)
hierfstat_data
fst_boot <- boot.ppfst(hierfstat_data, nboot = 10000, diploid=TRUE, quant=c(0.025,0.975))
fst_boot

```

## Comparing optitype vs HLA-LA

```{r}

## Avoiding this for a matter of time
pacman::p_load(tidyverse, midasHLA, styler, lintr, DT, irr)
#hla_la 2 fields
comp_hlala = htlv_df |>
  filter(country=="Peru")

# hla opt
htlv_opt = read_tsv("250304_HLAI/concatenated_HLA.txt", col_names = c("none", "A_1", "A_2", "B_1", "B_2", "C_1", "C_2", "none_reads", "none_obj", "ID")) |>
  dplyr::select(-starts_with("none")) |>
  dplyr::filter(!grepl("ATLSkin|1339", ID)) |>
  dplyr::filter(!grepl(serialsamples, ID)) |>
  mutate(
    ID=stringr::str_replace(ID, "^Peru_", "")
    ) 

comp_opt <- htlv_opt %>%
  mutate(
    HLA_A2 = pmap_chr(list(A_1, A_2), ~ {
      alleles <- c(...)
      paste(sort(alleles), collapse = "/")
    }),
    HLA_B2 = pmap_chr(list(B_1, B_2), ~ {
      alleles <- c(...)
      paste(sort(alleles), collapse = "/")
    }),
    HLA_C2 = pmap_chr(list(C_1, C_2), ~ {
      alleles <- c(...)
      paste(sort(alleles), collapse = "/")
    })
  ) |>
  dplyr::select(ID, HLA_A2, HLA_B2, HLA_C2)
comp_opt
comp_hlala = comp_hlala |>
  left_join(comp_opt, by="ID")

comp_hlala = comp_hlala |>
  filter(!is.na(HLA_A2))

# For HLA-A
kappa2(data.frame(comp_hlala$HLA_A, comp_hlala$HLA_A2))
kappa2(data.frame(comp_hlala$HLA_B, comp_hlala$HLA_B2))
kappa2(data.frame(comp_hlala$HLA_C, comp_hlala$HLA_C2))


mean(comp_hlala$HLA_A == comp_hlala$HLA_A2)* 100
mean(comp_hlala$HLA_B == comp_hlala$HLA_B2)* 100
mean(comp_hlala$HLA_C == comp_hlala$HLA_C2)* 100


```

## HLA divergence including all HLA-LA from Japan

```{ws{r}
#---packages
pacman::p_load(tidyverse, DT, tidyplots, httpgd, languageserver, midasHLA, stringr)


#---data loading
htlv_hla_la <- read_tsv("250304_HLAI/hla-la/v2/hla_la_concatenated_output.txt") #|>
    dplyr::filter(Q1 > .8)
DT::datatable(htlv_hla_la)
#---data wrangling
hlala_dataset <- htlv_hla_la |>
    filter(Locus %in% c("A", "B", "C")) |>
    dplyr::select(Locus, Chromosome, Allele, Sample) |>
    mutate(Allele = stringr::str_replace(Allele, "G", "")) |>
    mutate(Allele = str_trunc(Allele, width = 7, "right", ellipsis = "")) |>
    mutate(Alleles = paste0(Locus, Chromosome)) |>
    pivot_wider(values_from = "Allele", names_from = "Alleles", id_cols = "Sample") |>
    write_tsv("250304_HLAI/data/2508_hla_la_htlv_peru_and_japan.tsv")

hlala_dataset
#===After reading the file with hla_hed.py
#---reading the output
hla_hed = read_tsv("250304_HLAI/data/2508_v2_hla_total_hed.tsv")
#---data wrangling to make a new variable for countries

hla_hed = hla_hed |>
  mutate(Sample=stringr::str_replace(Sample, ".*_SS", "SS"))
hla_hed



hla_hed = hla_hed |>
    mutate(country=ifelse(str_detect(Sample, "IRID|SS"), "Peru", "Japan")) |>
    dplyr::filter(!grepl(serialsamples, Sample))

newdb = googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/14eZCI8lRsfIebXu-tpAxbuPEOhE6moAnTDKKKUQyNOI/edit?gid=1412798579#gid=1412798579", sheet="Sheet3")

### 1 events of LOH or complete homozygotus for HLA-1

hla_hed2 =hla_hed |>
    filter((HED_A+HED_B+HED_C)!=0) 

DT::datatable(hla_hed2)
hla_hed2


hla_hed2 = newdb |>
  select(HED_A, HED_B, HED_C, Mean_HE, strongy_lifetime, ID2) |>
  pivot_longer(-c(ID2, strongy_lifetime), names_to="HLAHED", values_to = "HED") |>
   mutate(
        HLAHED=HLAHED |> fct_recode(
            "Mean HLA-I"="Mean_HE",
            "HLA-A"="HED_A",
            "HLA-B"="HED_B",
            "HLA-C"="HED_C"
            ) |>
            fct_relevel(
                "Mean HLA-I",
                "HLA-A",
                "HLA-B",
                "HLA-C"
            ),
        strongy_lifetime=strongy_lifetime |> fct_recode("LSI"="YES", "non-LSI"="NO")
        ) 

hla_hed2 |>       
        tidyplot(x=strongy_lifetime, 
        y=HED, 
        color=strongy_lifetime) |>
    add_data_points_jitter() |>
    add_test_pvalue(
        hide_info = TRUE,
        label.size = 6/ggplot2::.pt
        ) |>
    add_violin() |>
    #remove_legend() |>
    adjust_x_axis(
        title="",
        labels = function(x) paste0(x, "\n(n = ", table(hla_hed2$strongy_lifetime)[x]/4, ")")
        ) |>
    adjust_y_axis(limits = c(0,NA), padding = c(0, NA)) |>
    adjust_title(face="bold") |>
        adjust_colors(
        new=c("#E69F00","#D55E00")
    ) |>
    tidyplots::remove_legend() |>
    split_plot(
        by = HLAHED, 
        ncol = 2, 
        nrow = 2,
        guides="keep")  |>
            save_plot(
        filename = "./output/2510_hla_la_LSI_HED.tiff",
        width = NA,
        height = NA,
        bg = "transparent"
    )

hla_hed2
hla_hed2  |>   
    pivot_longer(
        -c(country, Sample), 
        names_to="HLAHED", 
        values_to="HED"
        ) |>
    mutate(
        HLAHED=HLAHED |> fct_recode(
            "Mean HLA-I"="Mean_HE",
            "HLA-A"="HED_A",
            "HLA-B"="HED_B",
            "HLA-C"="HED_C"
            ) |>
            fct_relevel(
                "Mean HLA-I",
                "HLA-A",
                "HLA-B",
                "HLA-C"
            )
        ) |>
        tidyplot(x=country, 
        y=HED, 
        color=country) |>
    add_data_points_jitter() |>
    add_test_pvalue(
        hide_info = TRUE,
        label.size = 6/ggplot2::.pt
        ) |>
    add_violin() |>
    #remove_legend() |>
    adjust_x_axis(
        title="",
        labels = function(x) paste0(x, "\n(n = ", table(hla_hed2$country)[x], ")")
        ) |>
    adjust_y_axis(limits = c(0,NA), padding = c(0, NA)) |>
    adjust_title(face="bold") |>
        adjust_colors(
        new=c("#E69F00","#D55E00")
    ) |>
    split_plot(
        by = HLAHED, 
        ncol = 4, 
        nrow = 1,
        guides="keep") #|>
    save_plot(
        filename = "./output/2503_hla_la_japan_vs_peru.png",
        width = NA,
        height = NA,
        bg = "transparent"
    )

newdb = googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/14eZCI8lRsfIebXu-tpAxbuPEOhE6moAnTDKKKUQyNOI/edit?gid=1412798579#gid=1412798579", sheet="Sheet3") |>
  left_join(hla_hed, by=c("ID2"="Sample"))
newdb |>
googlesheets4::write_sheet("https://docs.google.com/spreadsheets/d/14eZCI8lRsfIebXu-tpAxbuPEOhE6moAnTDKKKUQyNOI/edit?gid=1412798579#gid=1412798579", sheet="Sheet3")
newdb = newdb|>
  mutate(logPVL=log(PVL+1))
cor_test <- cor.test(newdb$logPVL, newdb$Mean_HE, method = "spearman", use = "complete.obs")
print(cor_test)
print(cor_test)
newdb |>
  mutate(logPVL=log(PVL+1)) |>
  tidyplot(x=Status, y=Mean_HE) |>
  add_data_points() |>
  add_test_pvalue()
newdb

newdb$status1 = ifelse(newdb$Status=="death", 1, 0)

logit_model <- glm(strongy_lifetime  ~  Sex + Age + Mean_HE + PVL, 
                   data = newdb, 
                   family = binomial)
colnames(newdb)                
logit_model |>
  tbl_regression(exponentiate = TRUE)
summary(logit_model)
exp(cbind(OR = coef(logit_model), confint(logit_model)))

logit_model |>
  tbl_regression(exponentiate = TRUE) |>
  as_gt() |>
  gtsave(filename = "2508_sup_deathregresion.docx", path="./output")

# Including lost
htlv_mdata$status2 = ifelse(htlv_mdata$Status=="Death", 1, 0)
logit_model2 <- glm(status2  ~ Age + Sex + strongy_lifetime + PVL, 
                   data = htlv_mdata, 
                   family = binomial)
summary(logit_model)
logit_model2 |>
  tbl_regression(exponentiate = TRUE)
exp(cbind(OR = coef(logit_model), confint(logit_model)))

```

## Analysis

```{r}
### Analysis by groups

hlala_dataset_v2 = hlala_dataset |>
  mutate(Sample=stringr::str_replace(Sample, ".*_SS", "SS")) |>
  left_join(newdb, by=c("Sample"="ID2"))


# Step2: figures by 
A24A2 = hlala_dataset_v2 |>
  mutate(
    A0201 = ifelse(grepl("A\\*02:01", A1) | grepl("A\\*02:01", A2), "A*02:01(+)", "A*02:01(-)"),
    A2402 = ifelse(grepl("A\\*24:02", A1) | grepl("A\\*24:02", A2),  "A*24:02(+)", "A*24:02(-)")) |>
    pivot_longer(cols=c(A0201, A2402), names_to="HLA", values_to="haplotype")|>
    mutate(HLA=recode(HLA, A0201 = "HLA-A*02:01", A2402="HLA-A*24:02"))
A24A2
# comparison of IS by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=PVL) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "PVL (%)") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/2508_A24A2_PVL.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

##### Not performed 
# comparison of OCI by A24 and A02
A24A2|>
    tidyplot(x=haplotype, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    split_plot(by=HLA, tag_level=NULL) |>
    save_plot("output/S2_A_2_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

########### Not performed yet
# B35 and OCI
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35:01", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=gini) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Oligoclonality Index(OCI)") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_OCI.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 


# B35 part B
fig_B35 = htlv_hla |>
  mutate(
    B35= ifelse(grepl("B\\*35", B_1) | grepl("B\\*35", B_2), "B*35(+)", "B*35(-)")) |>
    tidyplot(x=B35, y=sum_freq) |>
    add_data_points_jitter(alpha=0.95) |>
    add_violin(draw_quantiles = 0.5, alpha=0.2) |>
    add_test_pvalue(method="t_test", method.args = list(alternative = "two.sided", var.equal = FALSE), hide_info = TRUE) |>
    adjust_y_axis(title = "Integration Sites (IS) \nNGS Abundance") |>
    adjust_x_axis(title = NULL) |>
    save_plot("output/S2_B35_vs_IS.png", view_plot = TRUE,
    bg = "transparent",
    width = 15,
    height = 8, 
    unit="cm") 

show(fig_B35)

```

## MIDAS

```{r}
#---Peru frequencies 
peru_hla = htlv_hla_la_wide |>
    filter(country=="Peru") |>
    mutate(ID=stringr::str_replace(ID, ".*_SS", "SS")) |>
    dplyr::select(-country) |>
    arrange(ID)
peru_hla

DT::datatable(peru_hla)
peru_hla |>
  write.csv("2508_HLA_forMIDAS.csv") 

##### from the backup
pacman::p_load(tidyverse, midasHLA)  
peru_hla = read.csv("2508_HLA_forMIDAS.csv") 

head(peru_hla)
peru_hla = peru_hla |>
  select(-X)

###. selection only by NGS
peru_hla
reduced_hla = peru_hla |>
  dplyr::select(c(1:7)) #, "DRB1_1", "DRB1_2"))

#---frequency calling
peru_hlacall <- reduceHlaCalls(reduced_hla, resolution = 4)
peru_hlacall
### metadata
pheno = newdb |>
  dplyr::select(ID2, status1, PVL, logPVL, strongy_lifetime, Status, disease_2) |>
  mutate(strongy=ifelse(strongy_lifetime=="YES", 1, 0),
  status2_lost=ifelse(Status %in% c("death", "LOST"), 1, 0),
  disease=ifelse(disease_2=="AC", 0, 1)) |>
  dplyr::select(-disease_2, -Status, -strongy_lifetime) |>
  mutate(ID2=
  case_when(
    ID2=="IRID062" ~"SS02",
    ID2=="IRID039" ~"SS04",
    ID2=="IRID015" ~"SS14",
    ID2=="IRID089" ~"SS20", 
    .default=ID2)) |>
  dplyr::rename(ID=ID2) |>
  arrange(ID)

pheno |>
  write.csv("2508_pheno_forMIDAS.csv")

pheno = read.csv("2508_pheno_forMIDAS.csv") |>
 select(-X)
head(pheno)
peru_hlacall
#___________________________________________________
?prepareMiDAS
#### preparing midas object
midas <- prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment =  c("hla_alleles", "hla_aa", "hla_g_groups", "hla_supertypes",
    "hla_NK_ligands", "hla_divergence", "hla_het"))

# ALLELES and cutoff >10%
midas <-
  filterByFrequency(
    object = midas,
    experiment = "hla_alleles",
    lower_frequency_cutoff = 0.10
    )

getHlaFrequencies(hla_calls = peru_hlacall, compare=FALSE) |>
filter(Freq>0.10) 


#### Preparing MIDAS
#midas <-
#  HWETest(
##    object = midas,
#    experiment = "hla_NK_ligands",
#    HWE_cutoff = 0.05 / 4,
#  as.MiDAS = TRUE
#    )


## Dominant Model Inheritance and Mortality Risk
object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_alleles", 
  inheritance_model = "dominant",
  lower_frequency_cutoff = 0.10, 
  upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )

results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I association with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_allele_Dominant.docx")

### AA + Dominant Model + Mortality Risk
midas <-
  filterByFrequency(
    object = midas,
    experiment = "hla_aa",
    lower_frequency_cutoff = 0.20
    )

object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_aa", 
  inheritance_model = "dominant",
  lower_frequency_cutoff = 0.10, 
  upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )
results

library(scales)
results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I association (Aminoacids) with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_aa_Dominant.docx")

## HLA G groups
peru_hlacall <- reduceHlaCalls(reduced_hla, resolution = 6)
midas <- prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment =  "hla_g_groups")



object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_g_groups", 
  inheritance_model = "dominant",
  #lower_frequency_cutoff = 0.10, 
  #upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )
results

library(scales)
results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I association (G-groups) with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_g-groups_Dominant.docx")

### HLA supertypes

peru_hlacall <- reduceHlaCalls(reduced_hla, resolution = 6)
midas <- prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment =  "hla_supertypes")

midas <-
  filterByFrequency(
    object = midas,
    experiment = "hla_supertypes",
    lower_frequency_cutoff = 0.10
    )

object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_supertypes", 
  inheritance_model = "recessive",
  #lower_frequency_cutoff = 0.10, 
  #upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )
results

library(scales)
results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I association (HLA-supertypes) with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_hla_supertypes_Dominant.docx")

## HLA NK ligands
peru_hlacall <- reduceHlaCalls(reduced_hla, resolution = 6)
midas <- prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment =  "hla_NK_ligands")

midas <-
  filterByFrequency(
    object = midas,
    experiment = "hla_supertypes",
    lower_frequency_cutoff = 0.10
    )

object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_NK_ligands", 
  inheritance_model = "dominant",
  lower_frequency_cutoff = 0.10, 
  upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )
results

library(scales)
results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I association (HLA-NK-Ligands) with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_hla_NK_ligands_Dominant.docx")


## HLA Divergence
peru_hlacall <- reduceHlaCalls(reduced_hla, resolution = 6)
midas <- prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment =  "hla_het")

midas <-
  filterByFrequency(
    object = midas,
    experiment = "hla_het",
    lower_frequency_cutoff = 0.10
    )

object <- glm(status1 ~ term , data = midas, family = binomial(link = "logit"))
results <-
  runMiDAS(
  object = object, 
  experiment = "hla_het", 
  #inheritance_model = "dominant",
  #lower_frequency_cutoff = 0.10, 
  #upper_frequency_cutoff = 0.95, 
  correction = "bonferroni", 
  #omnibus=TRUE,
  exponentiate = TRUE
    )
results

library(scales)
results |>
  as.data.frame() |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  regulartable() |>
  autofit() |>
  add_header_lines("TableX. HLA-I Heterozygozity with Mortality Risk (Dominant Inheritance Model)") |>
  flextable::save_as_docx(path="2508_hla_heterozygozyty_Dominant.docx")



library(knitr)
library(flextable)
regulartable(results) #|>
  kableExtra::save_kable("output/2508_hlaNK_deathrisk.docx")
### Running midas
library(MASS)
robust_model <- lm(PVL ~ term, data = midas) 
r1 =robust_model |>
tbl_regression()
¡
pheno
HLA_model <- glm(status2_lost ~ term, data = midas , family = binomial())
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_aa", 
  inheritance_model = "dominant",
  lower_frequency_cutoff = 0.05, 
  upper_frequency_cutoff = 0.95, 
  correction = "BH", 
  exponentiate = TRUE,
  omnibus=TRUE,
  omnibus_groups_filter =c("A_24", "A_68")
)

HLA_results
# Logistic regression
object <- glm(status2_lost ~ term + PVL, data = midas, family = binomial(link = "logit"))

results <-
  runMiDAS(
    object = object,
    experiment = "hla_alleles",
    inheritance_model = "dominant",
    conditional = FALSE,
    omnibus = FALSE,
    lower_frequency_cutoff = 0.05
    )
results |>
arrange(p.value)
kableResults(results)
############ NK Ligands
# Install development version from GitHub
devtools::install_github("Genentech/MiDAS")

prepareMiDAS(
  hla_calls = peru_hlacall,
  colData = pheno,
  experiment = "hla_aa")
pheno
NKlig_model <- glm(status2_lost ~ term, data = NKlig, family = binomial())
NKlig_results <- runMiDAS(
  NKlig_model,
  experiment = "hla_divergence",
  inheritance_model = "dominant",
  correction = "bonferroni",
  exponentiate = TRUE
)
NKlig_results


midas
custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome, conditional, omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
  HWE_cutoff = 0.05 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
  as.MiDAS = TRUE)

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    conditional=conditional,
    omnibus=omnibus,
    omnibus_groups_filter=additional
)
  return(HLA_results)
}

custom_midas(peru_hlacall, pheno_df, exp, inheritance,outcome, conditional, omnibus, additional)
# metadata
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) #|>
  #dplyr::filter(!is.na(gini))
colhtlv$Status
# Adjust data if it includes 0s or 1s
#colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_df = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

custom_midas <- function(hla_df, pheno_df, exp, inheritance,outcome)

# Define the models
models <- c("dominant", "recessive", "overdominant", "additive")
names(models) <- models  # Assign names to the models

# Initialize an empty list to store results
results_status <- list()
# Loop through the models
for (i in seq_along(models)) {
  # Apply custom_midas
  result <- custom_midas(
    hla_df = hla_df, 
    pheno_df = colhtlv, 
    exp = "hla_NK_ligands", 
    inheritance = models[i], 
    outcome = "Status",
    conditional=FALSE, #all results were not independent
    omnibus=FALSE,
    additional=NULL
  )
  
  # Add a column with the model name
  result$model <- names(models)[i]
  
  # Store the result in the list
  results_status[[i]] <- result
  combined_df <- do.call(rbind, results_status)
  return(combined_df)
}

datatable(combined_df)
write_tsv(combined_df, "output/2503_HLAanalysis_bydisease.tsv")  




custom_midas_het <- function(hla_df, pheno_df, exp, outcome){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = c("hla_het", "hla_divergence"))

  HLA_model <- glm(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA,
    family = binomial()
  )
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    exponentiate = TRUE
)
  return(HLA_results)
}

result <- custom_midas_het(
    hla_df = hla_df, 
    pheno_df = colhtlv,
    outcome="disease_2",
    exp = "hla_divergence") 
result

```

```{r}
# FOR IS PVL 
custom_midas_pvl <- function(hla_df, pheno_df, exp, inheritance,outcome,omnibus){ #, conditional) #omnibus, additional){
  # Prepare midasHLA
  HLA <- prepareMiDAS(
    hla_calls = hla_df,
    colData = pheno_df,
    experiment = exp)
  # Filter deviations from Hardy-Weinberg-Equilibrium (HWE) 
  HLA <- HWETest(
    object = HLA,
    experiment = exp,
    HWE_cutoff = 0.10 / 14, # p-value 0.05, working with only 14 alleles > 0.5%
    as.MiDAS = TRUE)

  HLA_model <- glm.nb(
    formula = as.formula(paste(outcome, "~ term")), # Dynamically create formula
    data = HLA)
    
  HLA_results <- runMiDAS(
    object = HLA_model, 
    experiment = exp, 
    inheritance_model = inheritance,
    lower_frequency_cutoff = 0.02, 
    upper_frequency_cutoff = 0.98, 
    correction = "bonferroni", 
    exponentiate = TRUE,
    #conditional=conditional,
    omnibus=omnibus,
    #omnibus_groups_filter=additional
)
  return(HLA_results)
}

##########
# data
# metadata
colhtlv2=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(sum_freq)) |>
  mutate(sum_freq=as.numeric(sum_freq))

hla_df2 = htlv_hla |>
  dplyr::filter(!is.na(sum_freq)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()
library(MASS)
models <- c("dominant", "recessive", "overdominant", "additive")
resultx = custom_midas_pvl(hla_df=hla_df2, pheno_df=colhtlv2, exp="hla_aa", inheritance="additive",outcome="sum_freq", omnibus=TRUE)
datatable(resultx)
```

```{r}
### ANALYSIS ASSOCIATION
#-- First, I am exploring the distribution of sum_freq (PVL) or gini index
#-----------------GINI INDEX (OCI) it is proportion (%)
htlv_hla <- htlv_hla[!is.na(htlv_hla$gini), ] #removing NAs

# Q-Q plot for normality (continuous data)
qqnorm(htlv_hla$gini)
qqline(htlv_hla$gini, col = "red") #it seems that it doesnt follow ND.

# Histogram of the response variable
hist(htlv_hla$gini, breaks = 30, main = "Distribution of Gini", xlab = "Gini")

"""
Beta regression model is designed for continuous data between 0 and 1. It uses a logit link function to model the mean of the Beta distribution
If the data is unimodal and symmetric, a Beta regression might be appropriate.
If the data is bimodal (peaks at 0 and 1), consider a zero-one-inflated Beta regression.
"""

# Simulate theoretical quantiles from the Beta distribution
fit_beta <- betareg(gini ~ 1, data = htlv_hla)
theoretical_quantiles <- qbeta(ppoints(length(htlv_hla$gini)), 
                               shape1 = fit_beta$coefficients$mean, 
                               shape2 = fit_beta$coefficients$precision)

# Create Q-Q plot
qqplot(theoretical_quantiles, htlv_hla$gini,
       main = "Q-Q Plot for Beta Distribution",
       xlab = "Theoretical Quantiles (Beta)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")
pacman::p_load(betareg)

# Corrected Q-Q plot for Poisson distribution
qqplot(qpois(ppoints(length(htlv_hla$gini)), lambda = mean(htlv_hla$gini)), 
       htlv_hla$gini,
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")



qqplot(qpois(ppoints(length(htlv_hla$sum_freq), lambda = mean(htlv_hla$sum_freq)), 
       main = "Q-Q Plot for Poisson Distribution",
       xlab = "Theoretical Quantiles (Poisson)",
       ylab = "Observed Quantiles")
abline(0, 1, col = "red")


 HLA = HWETest(
  object=midas,
  HWE_cutoff = 0.05 / 53,
  experiment = "hla_NK_ligands",
  as.MiDAS=TRUE)

# gini index as outcomd
colhtlv=htlv_hla |>
  select(ID, disease_2, strongy_lifetime, sum_freq, gini, Status) |>
  mutate(
    strongy_lifetime=as.factor(ifelse(strongy_lifetime=="Strongyloides (+)","1", "0")),
    disease_2=as.factor(ifelse(disease_2=="Asymptomatic carriers", "0", "1")),
    Status=as.factor(case_when(
      Status == "Death" ~ "1", 
      .default= "0"))
  ) |>
  dplyr::filter(!is.na(gini))

# Adjust data if it includes 0s or 1s
colhtlv$gini <- (colhtlv$gini * (length(colhtlv$gini) - 1) + 0.5) / length(colhtlv$gini)

hla_gini = htlv_hla |>
  dplyr::filter(!is.na(gini)) |>
  select(ID, A_1, A_2, B_1, B_2, C_1, C_2) |>
  as.data.frame()

# HLA divergence
midas <- prepareMiDAS(
  hla_calls = hla_gini,
  colData = colhtlv,
  experiment = "hla_divergence")

# Linear model
HLA_model <- lm(gini ~ term, data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence", 
  exponentiate = TRUE
)
# Beta reg model 
HLA_model <- gam(gini ~ term, family = betar(), data = midas)
HLA_results <- runMiDAS(
  object = HLA_model, 
  experiment = "hla_divergence",
  inheritance_model=""
)

HLA_results

dispersion <- sum(residuals(HLA_model, type = "pearson")^2) / df.residual(HLA_model)
dispersion # much greater than 1, the data is overdispersed.



HLA
freq |>
  filter(Counts>10)
midas_filtered_freq <-
  filterByFrequency(
    object = midas,
    experiment = "hla_alleles",
    lower_frequency_cutoff = 0.01
    )
midas_filtered_freq


object<-lm(gini~term, data=midas)
results <-
  runMiDAS(
    object = object,
    experiment = "hla_alleles",
    inheritance_model = "dominant",
    conditional = FALSE,
    omnibus = FALSE,
    lower_frequency_cutoff = 0.05
    )
kableResults(results)

htlv_clonality
results |>
  filter(p.value<0.05)

htlv_hla
```

https://github.com/cancerit/dockstore-cgpmap/wiki/1.-Reference-files

ftp://ftp.sanger.ac.uk/pub/cancer/dockstore/human/GRCh38_hla_decoy_ebv/bwa_idx_GRCh38_hla_decoy_ebv_bwamem2.tar.gz https://genentech.github.io/midasHLA/articles/MiDAS_vignette.html

https://hla.or.jp/med/frequency_search/en/allele/

## HLA-I or MHC-I

are divided into: - Classical HLA-I: A, B, C. Immunosurveillance and cancer immunotherapy. 60-90% in cancer are affected. (immunescape) a-heavy chain (HLA-A, HLA-B, HLA-C) b-2-microglobulin.

After ubiquitinylation, proteins degraded by immunoproteasome. Transportation by the transporter associated with antigen processing (TAP1/TAP2) into the ER. Tapasin + chaperones (calreticulin, ERp57) - Non-classical HLA-I:

### 

```{r}
library(tidyplots, DT, ggplot2)

plot_OR <- read_tsv("./output/2503_HLAanalysis_bydeath.tsv") |>
    dplyr::filter(p.value < 0.05) |>
    rename("CI_lower" = "conf.low", "CI_upper" = "conf.high") #|>

plot1 <- plot_OR |>
    tidyplot(x = estimate, y = allele) |>
    add_data_points() |>
    add(ggplot2::geom_errorbar(aes(xmin = CI_lower, xmax = CI_upper), width = .3, position = position_dodge(.9))) |>
    adjust_x_axis(
        title = "OR-adjusted to colinearity-LD",
        # limits=c(-5, 8)
    ) |>
    adjust_y_axis(title = "Alleles with p<0.05") |>
    add_annotation_line(x = 1, xend = 1.1, y = 0, yend = Inf, color = "grey") |>
    remove_legend() #|>

save_plot(plot1,
    filename = "./output/ORdeath.png",
    width = NA,
    height = NA,
    bg = "transparent"
)

datatable(plot_OR)
```

Using optitype, pairwise divergences of HLA-A, HLA-B, and HLA-C were similar

S20 seems to have a lower Quality for B

HED scores were calculated using a python script implemented in HLA-HED. \[https://github.com/sunhuaibo/HLA-HED\]

```{python}
## hed to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -o data/2503_hla_la_hed.txt

## paired to execute in the terminal
python hla_hed.py -d ./grantham_matrix.txt -f ./ABC_prot.fa -i ./data/2503_hla_la_htlv.tsv -p -o data/2503_hla_la_paired.txt
```

Background The antigen presentation from HLA-I represent the most effective control system to mantain HTLV-1 infection from HTLV-1 products THe spectrum of antigens presented by HLA type \## Heatmap of HLA-I evolutionary distances (HED) from Peruvian HTLV-1 cases

The heatmap shows z-score-normalized HED across HLA-I subtypes in HTLV-1 cases from Peru. The blue to red gradient indicates low to high HED between allele pairs, respectively.

```{r}
#---Supplementary heatmap distances
pairs1 <- read_tsv("data/2503_hla_la_paired.txt") |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    )
#---Create reverse pairs
reverse_pairs <- pairs1 |>
    mutate(Allele1A = Allele2, Allele2A = Allele1) |>
    filter(!paste(Allele1A, Allele2A) %in% paste(pairs1$Allele1, pairs1$Allele2)) |>
    mutate(Allele1 = Allele1A, Allele2 = Allele2A) |>
    select(-Allele1A, -Allele2A)
#---Combine original data and reverse pairs
complete_data <- bind_rows(pairs1, reverse_pairs) |>
    mutate(
        A1 = str_trunc(Allele1, width = 1, "right", ellipsis = ""),
        A2 = str_trunc(Allele2, width = 1, "right", ellipsis = "")
    ) |>
    mutate(
        A1 = recode(A1,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        ),
        A2 = recode(A2,
            "A" = "HLA-A",
            "B" = "HLA-B",
            "C" = "HLA-C"
        )
    ) |>
    arrange(Allele1, Allele2)
#---Heatmap for supplementary data
complete_data |>
    tidyplot(x = Allele1, y = Allele2, color = HED) |>
    add_heatmap() |>
    adjust_colors(colors_diverging_spectral) |>
    add(ggplot2::facet_grid(rows = vars(A2), cols = vars(A1), scales = "free", space = "free", margins = FALSE, switch = "both")) |>
    reverse_y_axis_labels() |>
    remove_x_axis_labels() |>
    remove_y_axis_labels() |>
    remove_x_axis_title() |>
    remove_y_axis_title() |>
    remove_x_axis_ticks() |>
    remove_y_axis_ticks() |>
    add(theme(panel.spacing = unit(0.05, "lines"))) |>
    save_plot(
        filename = "./output/2503_heatmap_hla.png",
        width = NA,
        height = NA,
        bg = "transparent"
    )
```

https://github.com/aiminy/IS-Seq-python3

https://github.com/nf-core/viralrecon

https://github.com/erickcastelli/HLA_genotyping

https://github.com/yshen25/HLA-Clus

https://www.youtube.com/watch?v=NNOEqMe4n74

https://grunwaldlab.github.io/Population_Genetics_in_R/

https://grunwaldlab.github.io/Population_Genetics_in_R/Data_Preparation.html

https://github.com/PapenfussLab/gridss

https://github.com/ICBI/viGEN

https://github.com/ClinicalGenomicsGBG/Viral_Integration_Pipeline

https://github.com/rosericazondekon/virusSeqPipeline

https://academic.oup.com/bioinformaticsadvances/article/4/1/vbae150/7815721

https://www.frontiersin.org/journals/microbiology/articles/10.3389/fmicb.2020.587306/full

https://github.com/GIGA-AnimalGenomics-BLV/Public/tree/master/PIC

https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-6647-4

https://github.com/ding-lab/VirusIntegrate https://github.com/SannaAb/Viral_Integration_Pipe

https://github.com/sara-javadzadeh/FastViFi

https://github.com/aehrc/isling

https://github.com/nf-core/viralintegration/blob/0.1.1/README.md

https://github.com/slowkow/hlabud

https://github.com/rbentham/HLAdivR/tree/48b2058f383ea75dad6e3c0f63bb1623ef5f0650

https://github.com/Genentech/midasHLA/blob/1a5bebfa67dd0f710a5ae5d4668fde9d3044e68a/NEWS.md?plain=1#L77

https://www.nature.com/articles/s41409-024-02341-z

https://jokergoo.github.io/ComplexHeatmap-reference/book/index.html